---
title: "Chapter 13 Session 2: Evaluating and Trusting a Multivariate Model"
subtitle: "Sections 13.7-13.9 plus assumptions for inference"
author: "Elementary Statistics"
format:
  revealjs:
    theme: simple
    slide-number: true
    chalkboard: true
    incremental: true
    code-fold: false
execute:
  echo: true
  warning: false
  message: false
---

```{r}
#| label: setup
#| include: false
library(coursekata)
library(dplyr)
library(ggplot2)

Neighborhood_model <- lm(PriceK ~ Neighborhood, data = Smallville)
HomeSizeK_model <- lm(PriceK ~ HomeSizeK, data = Smallville)
multi_model <- lm(PriceK ~ Neighborhood + HomeSizeK, data = Smallville)
```

## Last time

We fit:

$$
\text{PriceK}_i = b_0 + b_1\text{NeighborhoodEastside}_i + b_2\text{HomeSizeK}_i + e_i
$$

Today we ask:

> Is the multivariate model better than the empty model in the DGP?

---

## Model fit recap

```{r}
#| label: model-fit-recap
supernova(multi_model)
```

Look at:

- SS Model: error reduced
- SS Error: error leftover
- PRE / $R^2$: proportion reduction in error
- F: signal relative to leftover noise

---

## Venn diagram logic

Think of total variation in `PriceK` as a circle.

A multivariate model can reduce error through:

- unique variation explained by `HomeSizeK`
- unique variation explained by `Neighborhood`
- shared variation explained by both
- leftover unexplained error

---

## Unique vs shared explanation

```{r}
#| label: anova-components
supernova(multi_model)
```

The predictor rows ask:

> What does this variable add, given the other variable is already in the model?

---

## The model comparison question

Complex model:

$$
Y_i = \beta_0 + \beta_1X_{1i} + \beta_2X_{2i} + \epsilon_i
$$

Simple model:

$$
Y_i = \beta_0 + \epsilon_i
$$

We compare the multivariate model to the empty model.

---

## The null model for both predictors

Under the empty model:

$$
\beta_1 = 0 \quad \text{and} \quad \beta_2 = 0
$$

The predictors do not explain the outcome in the DGP.

---

## Simulating the empty model

To mimic no relationship, shuffle the outcome:

```{r}
#| label: one-shuffled-f
f(shuffle(PriceK) ~ Neighborhood + HomeSizeK, data = Smallville)
```

This keeps the predictors but breaks their relationship with price.

---

## Observed sample F

```{r}
#| label: observed-f
sample_f <- f(PriceK ~ Neighborhood + HomeSizeK, data = Smallville)
sample_f
```

F compares error reduced by the model against error left over.

---

## Build a null distribution of F

```{r}
#| label: null-f-distribution
set.seed(2026)
sdof <- do(1000) * f(shuffle(PriceK) ~ Neighborhood + HomeSizeK, data = Smallville)
head(sdof)
```

---

## Visualize the null distribution

```{r}
#| label: plot-null-f
#| fig-width: 7
#| fig-height: 4
gf_histogram(~ f, data = sdof, bins = 30) %>%
  gf_vline(xintercept = sample_f)
```

**Prompt:** Is the observed F typical under the empty model?

---

## Estimate the p-value

```{r}
#| label: p-value-from-simulation
tally(~ f > sample_f, data = sdof, format = "proportion")
```

This asks: how often did random shuffling produce an F as large as the observed F?

---

## Compare with the ANOVA table

```{r}
#| label: p-value-anova
supernova(multi_model)
```

The **Model** row tests the whole multivariate model against the empty model.

---

## Predictor rows are different questions

```{r}
#| label: predictor-rows
supernova(multi_model)
```

- Model row: Do the predictors jointly explain variation?
- `Neighborhood`: What does neighborhood add after home size?
- `HomeSizeK`: What does home size add after neighborhood?

---

## Confidence intervals for parameters

```{r}
#| label: confidence-intervals
confint(multi_model)
```

A confidence interval gives plausible values for each population parameter.

---

## Interpreting the CI for neighborhood

```{r}
#| label: ci-neighborhood
confint(multi_model)["NeighborhoodEastside", ]
```

Interpret as a plausible range for the Eastside-Downtown difference **holding home size constant**.

---

## Interpreting the CI for home size

```{r}
#| label: ci-homesize
confint(multi_model)["HomeSizeK", ]
```

Interpret as a plausible range for the price increase per additional 1000 square feet **holding neighborhood constant**.

---

## Theoretical inference

The model table gives theoretical tests based on the F distribution.

For each coefficient, theoretical inference relies on standard errors and t/F statistics.

```{r}
#| label: summary-model
summary(multi_model)
```

---

## Assumptions for inference

For theoretical inference in linear regression, we need to check whether the model is reasonable:

- linearity/additivity
- independence
- constant variance
- approximately normal residuals
- no extreme influential observations

---

## Residual plot: linearity and equal spread

```{r}
#| label: residual-plot
#| fig-width: 7
#| fig-height: 4
Smallville$fitted <- fitted(multi_model)
Smallville$resid <- resid(multi_model)

gf_point(resid ~ fitted, data = Smallville) %>%
  gf_hline(yintercept = 0)
```

Look for no obvious curve and roughly similar vertical spread.

---

## Residual normality check

```{r}
#| label: qqplot-residuals
#| fig-width: 6
#| fig-height: 4
qqnorm(resid(multi_model))
qqline(resid(multi_model))
```

This is most important for small samples and interval/test accuracy.

---

## Related predictors make interpretation tricky

When predictors are related, coefficients are still interpreted **holding other predictors constant**.

But this can become harder when there are few comparable observations.

Example: if Downtown homes tend to be larger, the model must separate size and neighborhood statistically.

---

## What can we conclude?

For this model, ask:

1. Does the whole model reduce error beyond the empty model?
2. What does each predictor add beyond the other?
3. Are assumptions reasonable enough for theoretical inference?
4. Are interpretations contextual and not causal unless design supports causality?

---

## Session 2 synthesis

Students should be able to:

- explain SS Model, SS Error, PRE, and F conceptually
- simulate a null distribution of F
- interpret the model-row p-value
- distinguish whole-model tests from predictor-specific tests
- interpret confidence intervals for multivariate coefficients
- check basic assumptions for regression inference
