---
title: "Chapter 13 Session 1: Building a Multivariate Model"
subtitle: "Sections 13.1-13.6: two predictors, coefficients, predictions, residuals, and SS"
author: "Elementary Statistics"
format:
  revealjs:
    theme: simple
    slide-number: true
    chalkboard: true
    incremental: true
    code-fold: false
execute:
  echo: true
  warning: false
  message: false
---

```{r}
#| label: setup
#| include: false
library(coursekata)
library(dplyr)
library(ggplot2)
```

## Today's question

Can we make better predictions by using **more than one explanatory variable**?

$$
\text{DATA} = \text{MODEL} + \text{ERROR}
$$

More predictors can reduce error, but interpretation gets more careful.

::: notes
Frame this as an extension, not a brand-new topic. The equation is still DATA = MODEL + ERROR.
:::

---

## The Smallville data

Outcome:

- `PriceK`: sale price in thousands of dollars

Predictors:

- `Neighborhood`: Downtown or Eastside
- `HomeSizeK`: square footage in thousands
- `HasFireplace`: fireplace indicator

```{r}
#| label: inspect-smallville
head(Smallville)
```

---

## Two single-predictor models

```{r}
#| label: single-predictor-models
Neighborhood_model <- lm(PriceK ~ Neighborhood, data = Smallville)
HomeSizeK_model <- lm(PriceK ~ HomeSizeK, data = Smallville)

supernova(Neighborhood_model)
supernova(HomeSizeK_model)
```

**Prompt:** Which model reduces more error?

---

## Visualizing each predictor

```{r}
#| label: visualize-one-predictor
#| fig-width: 7
#| fig-height: 4
gf_point(PriceK ~ HomeSizeK, data = Smallville) %>%
  gf_model(HomeSizeK_model)
```

---

## Add neighborhood visually

```{r}
#| label: color-neighborhood
#| fig-width: 7
#| fig-height: 4
gf_point(PriceK ~ HomeSizeK, data = Smallville,
         color = ~ Neighborhood) %>%
  gf_model(HomeSizeK_model, color = "black")
```

**Look for:** Are residuals patterned by neighborhood?

---

## Faceting by neighborhood

```{r}
#| label: facet-neighborhood
#| fig-width: 8
#| fig-height: 4
gf_point(PriceK ~ HomeSizeK, data = Smallville) %>%
  gf_facet_grid(. ~ Neighborhood)
```

**Prompt:** For homes of similar size, do prices still differ by neighborhood?

---

## Specifying a multivariate model

$$
Y_i = b_0 + b_1X_{1i} + b_2X_{2i} + e_i
$$

For Smallville:

$$
\text{PriceK}_i = b_0 + b_1\text{Neighborhood}_i + b_2\text{HomeSizeK}_i + e_i
$$

---

## Fitting the model in R

```{r}
#| label: fit-multivariate
multi_model <- lm(PriceK ~ Neighborhood + HomeSizeK, data = Smallville)
multi_model
```

Best-fitting estimates:

- choose values that minimize **SS Error**
- extend least squares to multiple predictors

---

## Write the fitted model

```{r}
#| label: coefficients
coef(multi_model)
```

$$
\widehat{\text{PriceK}} = b_0 + b_1(\text{NeighborhoodEastside}) + b_2(\text{HomeSizeK})
$$

**Important:** `NeighborhoodEastside = 1` for Eastside and `0` for Downtown.

---

## Interpreting the intercept

The intercept is the predicted price when:

- `NeighborhoodEastside = 0` → Downtown
- `HomeSizeK = 0`

So it is the model's Downtown baseline at 0K square feet.

**Question:** Is this substantively meaningful, or mainly mathematical?

---

## Interpreting `NeighborhoodEastside`

`NeighborhoodEastside` is the predicted difference between Eastside and Downtown homes...

**holding home size constant.**

```{r}
#| label: neighborhood-coef
coef(multi_model)["NeighborhoodEastside"]
```

Interpret in thousands of dollars.

---

## Interpreting `HomeSizeK`

`HomeSizeK` is the predicted change in price for each additional 1000 square feet...

**holding neighborhood constant.**

```{r}
#| label: homesize-coef
coef(multi_model)["HomeSizeK"]
```

---

## The hard phrase

> Holding the other variable constant

This means we compare observations that are the same on the other predictor.

- Same home size, different neighborhood
- Same neighborhood, different home size

---

## Compare single vs multivariate slope

```{r}
#| label: compare-slopes
coef(HomeSizeK_model)
coef(multi_model)
```

**Prompt:** Why does the `HomeSizeK` coefficient get smaller after adding `Neighborhood`?

---

## Predictions from the multivariate model

```{r}
#| label: predictions
Smallville$multi_predict <- predict(multi_model)
Smallville$multi_resid <- resid(multi_model)

Smallville %>%
  select(PriceK, Neighborhood, HomeSizeK, multi_predict, multi_resid) %>%
  head()
```

---

## Visualizing predictions

```{r}
#| label: prediction-triangles
#| fig-width: 7
#| fig-height: 4
gf_point(PriceK ~ HomeSizeK, color = ~ Neighborhood, data = Smallville) %>%
  gf_point(multi_predict ~ HomeSizeK, data = Smallville,
           color = "black", shape = 2)
```

The triangles are the model predictions.

---

## Parallel lines

```{r}
#| label: parallel-lines
#| fig-width: 7
#| fig-height: 4
gf_point(PriceK ~ HomeSizeK, color = ~ Neighborhood, data = Smallville) %>%
  gf_model(multi_model)
```

This additive model gives two lines with the **same slope** and different intercepts.

---

## Residuals still mean the same thing

$$
\text{Residual} = \text{Observed} - \text{Predicted}
$$

```{r}
#| label: residual-example
Smallville %>%
  mutate(check = multi_predict + multi_resid) %>%
  select(PriceK, multi_predict, multi_resid, check) %>%
  head()
```

---

## Sum of squared residuals

```{r}
#| label: ss-error
sum(resid(multi_model)^2)
supernova(multi_model)
```

SS Error tells us how much error is left after using both predictors.

---

## Comparing models by error reduction

```{r}
#| label: compare-model-fit
supernova(Neighborhood_model)
supernova(HomeSizeK_model)
supernova(multi_model)
```

**Prompt:** How much error does the multivariate model reduce compared with the empty model?

---

## Session 1 synthesis

By the end of today, students should be able to:

- explain why multiple predictors can improve prediction
- fit `lm(Y ~ X1 + X2)`
- interpret coefficients **holding the other predictor constant**
- calculate and interpret predictions and residuals
- connect residuals to SS Error and PRE

---

## Bridge to Session 2

Today: **What does the model say?**

Next time: **Could this model have arisen by random chance?**
