---
title: "CK Approach: Sternegth and Opportunities"
format: pptx
editor: source
---

```{r}
suppressPackageStartupMessages({
  library(tidyverse)
  library(patchwork)
  library(ggformula)
  library(mosaic)
})

theme_set(
  theme_bw() +
    theme(
    axis.text  = element_text(size = 20),
    axis.title = element_text(size = 22),
    legend.text  = element_text(size = 20),
    legend.title = element_text(size = 22)
  )
)

yaxis_off<- theme(
    axis.text.y  = element_blank(),  # remove y-axis tick labels
    axis.ticks.y = element_blank(),  # remove y-axis tick marks
    axis.title.y = element_blank()   # remove y-axis title
  )

source('../presentations/assets/utils.R')


gini <- function(x, normalize=TRUE){
  x<- as.factor(x)
  p <- table(x) |> prop.table()
  if(length(p)==1) return(0) else
    sum(p*(1-p)) * if(normalize) 1/(1-1/length(p)) else 1
}
entropy <- function(x, normalize=TRUE){
  x<- as.factor(x)
  p <- table(x) |> prop.table()
  if(length(p)==1) return(0) else
    -sum(p*log(p), na.rm = TRUE) * if(normalize) 1/log(length(p)) else 1
}

```

```{r}
x<- mtcars |> 
  mutate(cylinder=factor(cyl, uqcyl<- unique(cyl), 
                    paste0(uqcyl, 'cyl')),
         transmission=factor(am, 0:1, c('manual', 'automatic')))

```

## Association: Change in Variability

```{r}
nmdl0<- lm(mpg~NULL, data=x)
ss0<- sum(nmdl0$residuals^2)
nmdlcyl<- lm(mpg~cylinder, data=x)
sscyl<- sum(nmdlcyl$residuals^2)
gf_jitter(mpg~'overall', data=x, width=.1) %>% 
  gf_model(nmdl0, color='red') -> p0 
gf_jitter(mpg~cylinder, data=x, width=.1, 
          color=~cylinder, show.legend = FALSE) %>% 
  gf_model(nmdlcyl, color='red') ->pcyl
(p0+labs(x='', title='mpg~NULL',
         subtitle = sprintf('SST=%0.0f', ss0)))+
  (pcyl+yaxis_off+
     labs(title='       mpg~cylinder',
          subtitle = sprintf('        SSE=%0.0f', sscyl)))+
  plot_layout(widths = c(.25, .75))
```

## Extending it to Categorical

```{r}
x %>% gf_props(~'overall', fill=~transmission, show.legend = FALSE) %>% 
  gf_segment(1/2+1/2~.5+1.5, color='blue') %>% 
  gf_labs(y='Proportion')-> p1
GIT<- gini(x$transmission, normalize = TRUE)

x %>% gf_bar(~cylinder, fill=~transmission, position = 'fill',
               show.legend = FALSE) %>% 
  gf_segment(1/2+1/2~x+xend, color='blue', inherit = FALSE,
             data=data.frame(x=c(.525, 1.525, 2.525), 
                             xend=c(1.475, 2.475, 3.475))) %>% 
  gf_labs(y='Proportion')-> p2
GIE<- split(x$transmission, f=x$cylinder) |>
  sapply(entropy, normalize=FALSE) |> mean()

(p1+labs(x='', title='transmission~NULL',
         subtitle = sprintf('GIT=%0.3f', GIT)))+
  (p2+yaxis_off+
     labs(title='                 transmission~cylinder',
          subtitle = sprintf('                 AGI=%0.3f',
                             GIE)))+
  plot_layout(widths = c(.25, .75))

```

## Simulation: Great for Understanding Variability

```{r}
set.seed(2026)
xbar<- mean(x$mpg)
sd_mpg<- replicate(100, sample(x$mpg, replace = TRUE) |> mean())
mi<- quantile(sd_mpg, c(0.025, 0.975))
me<- diff(mi)/2
ci<-sort(2*xbar-mi)
# t - conf-int
s<- sd(x$mpg); n<- length(x$mpg); se<- s/sqrt(n); tstr<- -qt(.025, n-1)
tci<- xbar+c(-1,1)*tstr*se

gf_histogram(~sd_mpg, bins = 10) |> 
  gf_labs(title=sprintf('Sampling Distribution of MPG (n=%g)',
                        nrow(x)), x='mpg') |>
  gf_vline(xintercept = mi, linetype = 2) |>
  gf_vline(xintercept = mean(x$mpg),color='red') + yaxis_off
  
```
## Employing Symmetry & Elevating Mathematical Rigor

```{r}
gf_histogram(~sd_mpg, bins = 10) |> 
  gf_labs(title=sprintf('Sampling Distribution of MPG (n=%g)',
                        nrow(x)), x='mpg') |>
  gf_vline(xintercept = mi, linetype = 2) |>
  gf_vline(xintercept = mean(x$mpg),color='red') |>
  gf_segment(1+1~cl+cu, inhirit=FALSE, color=~"Symmetric CI",
             size=1.5,
             data = data.frame(cl=xbar-me, cu=xbar+me),
             arrow = grid::arrow(ends = 'both',
                                 type = 'closed',
                                 length = unit(.1, 'inches'))
  ) |>
  gf_segment(0+0~cl+cu, inhirit=FALSE, color=~"Solving for μ",
             data = data.frame(cl=ci[1], cu=ci[2]),
             size=1.5,
             arrow = grid::arrow(ends = 'both',
                                 type = 'closed',
                                 length = unit(.1, 'inches'))
  ) |>
  gf_segment(2+2~cl+cu, inhirit=FALSE, color=~"t-Conf.Int.",
             data = data.frame(cl=tci[1], cu=tci[2]),
             size=1.5,
             arrow = grid::arrow(ends = 'both',
                                 type = 'closed',
                                 length = unit(.1, 'inches'))
  ) |>
  gf_refine(
    scale_color_manual(
      name = "Confidence Intervals",          # Title of the legend
      values = c("t-Conf.Int."='orange',
                 "Solving for μ" = "red",
                 "Symmetric CI" = "purple"), # Custom colors
      breaks = c("t-Conf.Int.", "Symmetric CI", "Solving for μ") # Controls the number of lines/items shown
    ),
    # Use theme() to shrink the legend components
    theme(
      legend.title = element_text(size = 12),       # Shrink title font
      legend.text = element_text(size = 10),        # Shrink item font
      legend.key.size = unit(.7, "cm"),           # Shrink the colored icon box
      legend.spacing.y = unit(0.02, "cm")          # Pack the rows closer together
    )
  ) + yaxis_off
  
```

## Simulation: Knowing the Limits

- Modeling Complex Situations:
  - Single Variable in a Multivariate Model

- Post Hoc Analysis
  - Tukey's Honest Significant Difference
  - Pairwise t-Test
  
- Environmental Footprint