Excellent! Now you want interactivity with sliders and dynamic graphs. With webr, you have several options:

## Option 1: Use Shiny-style inputs with `ojs` (Observable JS)

Quarto supports Observable JS which works great with webr for interactive controls:

```markdown
---
format:
  live-revealjs:
    slide-number: true
---

```{ojs}
//| echo: false
viewof n = Inputs.range([10, 1000], {step: 10, label: "Sample size:", value: 100})
```

```{webr-r}
#| autorun: true
# Access the ojs value
n <- as.integer(ojs_define(n = n)$n)
x <- rnorm(n)
hist(x, main = paste("Histogram of", n, "random normals"))
```
```

## Option 2: Use htmlwidgets packages in webr

Many R htmlwidgets work in webr:

```markdown
::: {.fragment}
```{webr-r}
library(plotly)

# Create interactive plot
plot_ly(mtcars, x = ~wt, y = ~mpg, type = 'scatter', mode = 'markers')
```
:::
```

## Option 3: Pure webr with HTML inputs

Create custom HTML controls:

```markdown
<input type="range" id="slider" min="10" max="100" value="50">
<span id="value">50</span>

```{webr-r}
#| context: interactive
library(htmltools)

# JavaScript to update R when slider changes
tags$script(HTML("
  document.getElementById('slider').addEventListener('input', function(e) {
    const val = e.target.value;
    document.getElementById('value').innerText = val;
    // Trigger R code with new value
    webr.evalR(`
      n <- ${val}
      x <- rnorm(n)
      hist(x, main = paste('n =', n))
    `);
  });
"))
```
```

## Option 4: Use `{crosstalk}` for linked views

```markdown
```{webr-r}
library(crosstalk)
library(plotly)

shared_mtcars <- SharedData$new(mtcars)

filter_slider("cyl", "Cylinders", shared_mtcars, ~cyl)
plot_ly(shared_mtcars, x = ~wt, y = ~mpg) %>% add_markers()
```
```

## Recommended approach for presentations:

For Revealjs + webr, I'd suggest **Option 1 (Observable JS)** because:
- Clean, simple syntax
- Works seamlessly with webr
- Professional-looking controls

Want me to show you a complete example slide with a specific type of interactive visualization you have in mind?

# R installation issues:

```bash
micromamba install -c conda-forge r-ggformula

```

```bash
micromamba install -n my_r_env libgit2 libcurl libxml2 openssl make gcc

```

```{r}
install.packages(c("rcmdcheck", "usethis", "remotes", "httr", "jsonlite", "git2r"))
install.packages('devtools')
devtools::install_github("ProjectMOSAIC/ggformula")
```

```bash
# on mac
brew install pkg-config cairo libpng freetype
#on linux
sudo apt-get update
sudo apt-get install libgtk2.0-dev libcairo2-dev libxt-dev libpng-dev libjpeg-dev libtiff5-dev libgif-dev
```

```bash
brew install cairo
brew install freetype
brew install harfbuzz
brew install fribidi
```

```{r}
# Install gdtools first
install.packages("gdtools")  # resolves font and graphics issues

# Install ggiraph next
install.packages("ggiraph")  

# Finally, devtools
install.packages("devtools")  

# Install ggformula from CRAN
install.packages("ggformula")
```


# New Search Results (2/27/2026)
You can share a value **from WebR → JavaScript/HTML** using the small bridge that WebR exposes: `webr::js_eval()` (or `webr::eval_js()` depending on version). The key idea is:

**Anything you assign to `globalThis` in JavaScript becomes visible to the HTML page.  
Anything you return from R via `webr::js_eval()` becomes a JavaScript value.**

To make this concrete, here are the three reliable patterns you can use in Quarto, revealjs, or any HTML page that embeds WebR.

---

## 🧩 1. WebR → JavaScript using `webr::js_eval()`

Inside a WebR chunk:

```r
#| engine: webr

x <- 42
webr::js_eval(sprintf("globalThis.fromR = %s;", x))
```

Now in your HTML/JS:

```js
console.log(globalThis.fromR);   // 42
```

This is the simplest and most direct way to expose an R value to the page.

---

## 🧩 2. WebR → JavaScript for vectors, lists, data frames

WebR automatically converts R objects to JS equivalents.

### R vector → JS array

```r
#| engine: webr
v <- c(10, 20, 30)
webr::js_eval(sprintf("globalThis.rVector = %s;", jsonlite::toJSON(v)))
```

Then in JS:

```js
console.log(globalThis.rVector);   // [10, 20, 30]
```

### R data frame → JS array of objects

```r
#| engine: webr
df <- data.frame(x = 1:3, y = c("a","b","c"))
webr::js_eval(sprintf("globalThis.rData = %s;", jsonlite::toJSON(df)))
```

Then in JS:

```js
console.log(globalThis.rData);
// [
//   { "x": 1, "y": "a" },
//   { "x": 2, "y": "b" },
//   { "x": 3, "y": "c" }
// ]
```

This is extremely useful for interactive dashboards.

---

## 🧩 3. WebR → HTML elements directly

You can write into the DOM from R by calling JS that manipulates HTML.

### HTML

```html
<div id="result"></div>
```

### WebR chunk

```r
#| engine: webr
value <- mean(c(1,2,3,4))

webr::js_eval(sprintf("
  document.getElementById('result').textContent =
    'Mean from R: %s';
", value))
```

This updates the HTML page live.

---

## 🧩 4. WebR → JS callback functions

You can even trigger JS functions defined in the page.

### JS in your HTML

```js
function updatePlotFromR(val) {
  console.log("R sent:", val);
}
```

### WebR chunk

```r
#| engine: webr
n <- 7
webr::js_eval(sprintf("updatePlotFromR(%s);", n))
```

This is the cleanest pattern for interactive teaching demos.

---

## 🧩 5. WebR → OJS (if you use Observable cells)

OJS can read any JS global:

```ojs
myValue = globalThis.fromR
```

So the pipeline becomes:

**WebR → JS global → OJS**

This is the mirror of the OJS → WebR pattern you already used.

---

## Summary of the bridge

| Direction | Method | Notes |
|----------|--------|-------|
| **WebR → JS** | `webr::js_eval("...")` | Most direct and reliable |
| **JS → HTML** | DOM manipulation | Use `document.getElementById()` etc. |
| **WebR → HTML** | R calls JS that updates DOM | Great for teaching |
| **WebR → OJS** | Write to `globalThis` | OJS can read globals |

---

If you want, I can show you a **minimal working Quarto example** where a WebR slider updates a JS chart, or where WebR computes a statistic and updates an HTML table live.
