Excellent! Now you want interactivity with sliders and dynamic graphs. With webr, you have several options:

## Option 1: Use Shiny-style inputs with `ojs` (Observable JS)

Quarto supports Observable JS which works great with webr for interactive controls:

```markdown
---
format:
  live-revealjs:
    slide-number: true
---

```{ojs}
//| echo: false
viewof n = Inputs.range([10, 1000], {step: 10, label: "Sample size:", value: 100})
```

```{webr-r}
#| autorun: true
# Access the ojs value
n <- as.integer(ojs_define(n = n)$n)
x <- rnorm(n)
hist(x, main = paste("Histogram of", n, "random normals"))
```
```

## Option 2: Use htmlwidgets packages in webr

Many R htmlwidgets work in webr:

```markdown
::: {.fragment}
```{webr-r}
library(plotly)

# Create interactive plot
plot_ly(mtcars, x = ~wt, y = ~mpg, type = 'scatter', mode = 'markers')
```
:::
```

## Option 3: Pure webr with HTML inputs

Create custom HTML controls:

```markdown
<input type="range" id="slider" min="10" max="100" value="50">
<span id="value">50</span>

```{webr-r}
#| context: interactive
library(htmltools)

# JavaScript to update R when slider changes
tags$script(HTML("
  document.getElementById('slider').addEventListener('input', function(e) {
    const val = e.target.value;
    document.getElementById('value').innerText = val;
    // Trigger R code with new value
    webr.evalR(`
      n <- ${val}
      x <- rnorm(n)
      hist(x, main = paste('n =', n))
    `);
  });
"))
```
```

## Option 4: Use `{crosstalk}` for linked views

```markdown
```{webr-r}
library(crosstalk)
library(plotly)

shared_mtcars <- SharedData$new(mtcars)

filter_slider("cyl", "Cylinders", shared_mtcars, ~cyl)
plot_ly(shared_mtcars, x = ~wt, y = ~mpg) %>% add_markers()
```
```

## Recommended approach for presentations:

For Revealjs + webr, I'd suggest **Option 1 (Observable JS)** because:
- Clean, simple syntax
- Works seamlessly with webr
- Professional-looking controls

Want me to show you a complete example slide with a specific type of interactive visualization you have in mind?

# R installation issues:

```bash
micromamba install -c conda-forge r-ggformula

```

```bash
micromamba install -n my_r_env libgit2 libcurl libxml2 openssl make gcc

```

```{r}
install.packages(c("rcmdcheck", "usethis", "remotes", "httr", "jsonlite", "git2r"))
install.packages('devtools')
devtools::install_github("ProjectMOSAIC/ggformula")
```

```bash
# on mac
brew install pkg-config cairo libpng freetype
#on linux
sudo apt-get update
sudo apt-get install libgtk2.0-dev libcairo2-dev libxt-dev libpng-dev libjpeg-dev libtiff5-dev libgif-dev
```

```bash
brew install cairo
brew install freetype
brew install harfbuzz
brew install fribidi
```

```{r}
# Install gdtools first
install.packages("gdtools")  # resolves font and graphics issues

# Install ggiraph next
install.packages("ggiraph")  

# Finally, devtools
install.packages("devtools")  

# Install ggformula from CRAN
install.packages("ggformula")
```
