---
title: "Sample Exam 2"
sidebar: false
page-navigation: false
bread-crumbs: false
format: 
  html:
    toc: true
    toc-float: true
    css:
      - ../css/h4h5.css
      - ../css/h3collapsable.css
---

```{r, echo=FALSE}
options(knitr.kable.NA = '')
knitr::opts_chunk$set(echo = FALSE)
```


### Part A: Conceptual (Multiple Choice)

#### What does a small p-value indicate?

##### The null hypothesis is true
##### Strong evidence against the null hypothesis
##### The sample size is small
##### The model has no error

---

#### What does R² measure?

##### The slope of the regression line
##### The variability in the predictor
##### The proportion of variability explained by the model
##### The p-value of the test

---

#### A Type I error occurs when:

##### You fail to reject a false null hypothesis
##### You reject a true null hypothesis
##### You accept a false alternative hypothesis
##### The sample size is too large

---

#### Which test is used to compare more than two group means?

##### t-test
##### ANOVA
##### Chi-square
##### Regression

---

#### In regression, what does b₁ represent?

##### The average of Y
##### The predicted value when X = 0
##### The average change in Y for a one-unit increase in X
##### The variance of Y

---

### Part B: Short Answer

#### Explain the difference between a simple and a complex model.

\

---

#### What is a null distribution and why is it important?

\

---

#### What does it mean if a result is statistically significant at α = 0.05?

\

---

### Part C: Interpretation

#### A regression analysis gives:

```{r, echo=FALSE, include=FALSE, eval=FALSE}
summary(lm((drat)~I(scale(wt)+scale(qsec)), data=mtcars))
```

| Terms | Estimate | St##### Error | t value | Pr(>|t|) |  
|-------------|---------|---------|--------|---------|
| (Intercept) | 3.59656 | 0.08410 | 42.765 | < 2e-16 |
| x  | -0.20124 | 0.06651 | [t] | 0.00505| 


Answer:

##### Interpret the slope

\

##### Find the missing `[t]`

\

##### State whether the result is significant

\

##### Write the equation for the regression (use $y$ as the outcome variable):

\

##### Find and interpret the value of $y$ when $x$ is 0.75.

\

---

#### An ANOVA test gives:

```{r, echo=FALSE, include=FALSE, eval=FALSE, results='asis', }
#supernova::supernova(lm(drat~factor(am+vs), data=mtcars))
anv.tbl<- 
  within(
    data.frame(
      SS=c(5.1, 3.7),
      df=c(4, 23),
      row.names = c('Model (error reduced)', 'Error (from model)')
    ),{
      MS<- SS/df
      F<- c(MS[1]/MS[2], NA)
      PRE<- c(SS[1]/sum(SS), NA)
      p<- 1-pf(F, df1 = df[1], df2 = df[2])
    })[, c('SS', 'df', 'MS', 'F', 'PRE', 'p')]
anv.tbl |> format(digits=2, width=6, nsmall=2, scientific=F) |> knitr::kable(format = 'pipe')
```

|                      |SS   |df    |MS   |F    |PRE  |p       |
|:---------------------|:----|:-----|:----|:----|:----|:-------|
|Model (error reduce##### |5.10 |4.00  | [a] | [F] | [R2] |0.0004 |
|Error (from model)    |3.70 |23.00 | [b] |    |    |       |

 

Answer:

##### Find the missing `[F]` value.

\

##### How many groups are being abalyzed?

\

##### What are the hypotheses this table is testing?


##### What is the conclusion of the test?

\

##### Find the missing `[R2]` and interpret it.

\

---

### Part D: Applied Thinking

#### A researcher runs 10 different tests at α = 0.05.

##### What problem might occur?

\

##### How could it be addressed?

\

---

#### You observe a strong relationship in a sample, but the p-value is large. Why might this happen?

\

---

### Part E: Data Reasoning

#### Given a contingency table (Gender × Job):

##### What test should be used?

\


##### What would the null hypothesis be?

\

---

#### The following summary of $y$ is given:

```{r, echo=FALSE}
aggregate(drat~factor(vs, 0:1, c('A', 'B')),
          data=mtcars, length) |> setNames(c('Group', 'n')) |>
  merge(aggregate(drat~factor(vs, 0:1, c('A', 'B')),
          data=mtcars, mean) |> setNames(c('Group', 'Mean'))) |>
  merge(aggregate(drat~factor(vs, 0:1, c('A', 'B')),
          data=mtcars, sd) |> setNames(c('Group', 'SD'))) |> 
  knitr::kable(digits = 2)
#with(mtcars, t.test(x = drat[vs==0], y=drat[vs==1], var.equal=TRUE)) 
```

##### What is the differnce of the sample means?

\

##### What is the standard error of the differnce of the sample means?

\

##### Compute $t=\frac{\bar x_B-\bar x_A}{SE}$.

\

##### what is the distribution of $t$, if we assume the group variances are equal?

\

##### Write a conclusion if the $p\text{-value}=0.012$ and $H_a: \mu_A\ne\mu_B$.

\

##### What is the Cohen's $d$ and what does it mean?

\

##### Write the `lm` equation for `Y~X` and the estimate values for $\beta_0$ and $\beta_1$?

\

---

### Challenge Questions

#### Explain how the t-distribution differs from the normal distribution.

\

#### Why do we use an F-test instead of multiple t-tests in ANOVA?

\

---