---
title: "Exam 2"
subtitle: "Statistics — Chapters 5–11"
date: today
format:
  html:
    theme: cosmo
    toc: false
    number-sections: false
    self-contained: true
    css: |
      body { max-width: 820px; margin: auto; font-size: 15px; }
      .question { margin-bottom: 2em; padding: 1em 1.25em; border: 1px solid #e0e0e0; border-radius: 8px; background: #fafafa; }
      .q-num { font-size: 0.75em; color: #888; margin-bottom: 0.3em; font-weight: 600; text-transform: uppercase; letter-spacing: 0.05em; }
      .section-title { font-size: 1em; font-weight: 600; color: #444; border-bottom: 1px solid #ddd; padding-bottom: 4px; margin: 2em 0 1em; }
      .short-answer textarea { width: 100%; min-height: 80px; border: 1px solid #ccc; border-radius: 6px; padding: 8px; font-size: 14px; font-family: inherit; }
      .long-answer textarea { width: 100%; min-height: 140px; border: 1px solid #ccc; border-radius: 6px; padding: 8px; font-size: 14px; font-family: inherit; }
      .instructions { background: #f0f4ff; border-left: 4px solid #4a6cf7; padding: 0.75em 1em; border-radius: 4px; margin-bottom: 2em; font-size: 0.93em; }
      table { border-collapse: collapse; width: 100%; margin: 0.75em 0; }
      th, td { border: 1px solid #ccc; padding: 6px 10px; text-align: center; }
      th { background: #f0f0f0; }
      .formula-box { background: #f8f8f8; border: 1px solid #ddd; border-radius: 6px; padding: 0.75em 1em; margin: 0.5em 0 1em; font-size: 0.9em; }
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

::: instructions
**Instructions:** This exam covers Chapters 5–11 of CourseKata. For multiple choice questions, circle or select the single best answer. For short answer questions, show your work where calculations are involved. You may use R. Total: 100 points. Time: 75 minutes.
:::

---

## Part 1: Models and Variability (20 pts)

::: {.section-title}
Questions 1–4 · 5 points each
:::

::: question
**Q1.** A researcher models students' final exam scores using only an intercept (empty model). The sample mean is $\bar{y} = 74$ and SST = 2400. Which of the following is the predicted value $\hat{y}_i$ for *every* student under the empty model?

a) 0  
b) 74  
c) The student's actual score  
d) The median score  

*Answer:* ______
:::

::: question
**Q2.** A model is fit to 30 observations with SSE = 360 and SST = 900. Which of the following statements is correct?

a) $R^2 = 0.40$; the model explains 40% of the variability in the outcome.  
b) $R^2 = 0.60$; the model explains 60% of the variability in the outcome.  
c) $R^2 = 2.50$; the model over-explains the variability.  
d) $R^2 = 0.60$; the model explains 60% of the residuals.  

*Answer:* ______
:::

::: question
**Q3.** Which of the following correctly classifies a model with one numerical predictor and one binary categorical predictor?

a) Empty model (1 parameter)  
b) Simple model (2 parameters)  
c) Complex model (3 or more parameters)  
d) It cannot be classified without more information  

*Answer:* ______
:::

::: question
**Q4.** If $\text{SST} = 500$, $\text{SSE} = 500$, what can you conclude?

a) The model perfectly predicts all outcomes.  
b) The model explains none of the variability — it performs no better than the empty model.  
c) $R^2 = 1$  
d) $R^2 = -1$  

*Answer:* ______
:::

---

## Part 2: Hypothesis Testing Framework (20 pts)

::: {.section-title}
Questions 5–8 · 5 points each
:::

::: question
**Q5.** A p-value of 0.12 is obtained at significance level $\alpha = 0.05$. What is the correct decision and conclusion?

a) Reject $H_0$; there is significant evidence for $H_a$.  
b) Fail to reject $H_0$; there is not enough evidence against $H_0$.  
c) Accept $H_0$; the null hypothesis is true.  
d) Reject $H_0$; the result is marginally significant.  

*Answer:* ______
:::

::: question
**Q6.** Which of the following describes a Type I error?

a) Failing to reject $H_0$ when $H_0$ is false.  
b) Rejecting $H_0$ when $H_0$ is true.  
c) Using a test statistic that does not match the data.  
d) Setting $\alpha$ too small.  

*Answer:* ______
:::

::: question
**Q7.** A simulation-based null distribution for the difference in group means $(\bar{y}_A - \bar{y}_B)$ is generated by:

a) Randomly drawing new samples from a normal distribution with mean $\beta_1$.  
b) Adding random noise to the observed group means.  
c) Shuffling all $y_i$ values and randomly reallocating $n_A$ to group A and $n_B$ to group B.  
d) Resampling with replacement from each group separately.  

*Answer:* ______
:::

::: question
**Q8.** A researcher runs 6 pairwise t-tests each at $\alpha = 0.05$ to compare group means. What is the primary concern?

a) Each test will have reduced power.  
b) The overall Type I error rate inflates above 0.05.  
c) The degrees of freedom will be too small.  
d) The t-distribution is no longer valid.  

*Answer:* ______
:::

---

## Part 3: Testing One Mean & Two Groups (20 pts)

::: {.section-title}
Questions 9–12 · 5 points each
:::

::: question
**Q9.** A sample of $n = 36$ students has $\bar{y} = 68$ and $s = 12$. You test $H_0: \mu = 65$ vs. $H_a: \mu \neq 65$. What is the observed t-statistic?

a) 0.25  
b) 1.50  
c) 3.00  
d) 0.50  

*Answer:* ______
:::

::: question
**Q10.** For a two-sample t-test assuming equal variances with $n_A = 18$ and $n_B = 22$, what are the degrees of freedom?

a) 17  
b) 21  
c) 38  
d) 40  

*Answer:* ______
:::

::: question
**Q11.** Group A has SS(A) = 240, $n_A = 13$. Group B has SS(B) = 320, $n_B = 17$. What is the pooled variance $s_p^2$?

a) 20.0  
b) 20.3  
c) 22.4  
d) 28.0  

*Answer:* ______
:::

::: question
**Q12.** The mathematical null distribution $t = \frac{\bar{y} - \beta_0}{s/\sqrt{n}}$ follows a t-distribution with $df = n - 1$ under which assumption(s)?

a) The population variance is known.  
b) The errors are normally distributed, OR $n$ is large.  
c) The observations are taken from a stratified random sample.  
d) Both groups have equal variance.  

*Answer:* ______
:::

---

## Part 4: Linear Regression (20 pts)

::: {.section-title}
Questions 13–16 · 5 points each
:::

::: question
**Q13.** The correlation between $x$ and $y$ is $r = 0.75$, $s_y = 10$, $s_x = 5$. What is the estimated least-squares slope $b_1$?

a) 0.375  
b) 0.75  
c) 1.50  
d) 3.75  

*Answer:* ______
:::

::: question
**Q14.** Given $b_1 = 1.5$, $\bar{y} = 40$, $\bar{x} = 20$, what is the intercept $b_0$?

a) 10  
b) 70  
c) 20  
d) -10  

*Answer:* ______
:::

::: question
**Q15.** A regression model has $n = 28$ observations. The error variance $s^2 = \text{SSE}/(n-2)$ has how many degrees of freedom?

a) 27  
b) 26  
c) 25  
d) 28  

*Answer:* ______
:::

::: question
**Q16.** To simulate the null distribution of slope $b_1$ under $H_0: \beta_1 = 0$, you should:

a) Resample pairs $(x_i, y_i)$ with replacement.  
b) Shuffle all $y_i$ values and pair each with a fixed $x_i$.  
c) Generate new $x$ values from a normal distribution.  
d) Add random noise to the observed slope.  

*Answer:* ______
:::

---

## Part 5: Complex Models & ANOVA (20 pts)

::: {.section-title}
Questions 17–20 · 5 points each
:::

::: question
**Q17.** An ANOVA model has $k = 3$ predictors and $n = 60$ observations. What are the degrees of freedom for the F-statistic?

a) $df_1 = 3$, $df_2 = 56$  
b) $df_1 = 2$, $df_2 = 57$  
c) $df_1 = 3$, $df_2 = 59$  
d) $df_1 = 4$, $df_2 = 56$  

*Answer:* ______
:::

::: question
**Q18.** With SSM = 480 and SSE = 840, $k = 3$, $n = 60$, what is the observed F-statistic?

a) 0.57  
b) 5.71  
c) 10.67  
d) 0.19  

*Answer:* ______
:::

::: question
**Q19.** A categorical variable has 5 levels. How many indicator (dummy) variables are needed to include it in a regression model?

a) 5  
b) 4  
c) 3  
d) 1  

*Answer:* ______
:::

::: question
**Q20.** After a significant omnibus F-test in an ANOVA with a categorical predictor at 4 levels, the appropriate next step is:

a) Stop; the F-test already identifies which groups differ.  
b) Run pairwise comparisons with a Bonferroni-corrected $\alpha$ or adjusted p-values.  
c) Re-run the model without the categorical predictor.  
d) Increase the sample size and re-test.  

*Answer:* ______
:::

---

## Part 6: Short Answer (30 pts)

::: {.section-title}
Questions 21–24 · point values noted
:::

::: question
**Q21. (6 pts)** A statistics student collects data on hours studied ($x$) and exam score ($y$) from 20 students. The output below is from a simple linear regression:

| | Estimate | Std. Error | t value | Pr(>\|t\|) |
|---|---|---|---|---|
| (Intercept) | 52.0 | 4.1 | 12.68 | <0.001 |
| hours | 3.8 | 1.2 | 3.17 | 0.005 |

**a)** Write the estimated regression equation. (1 pt)

::: short-answer
<textarea placeholder="ŷ = ..."></textarea>
:::

**b)** Interpret the slope $b_1 = 3.8$ in context. (2 pts)

::: short-answer
<textarea placeholder="For each additional hour studied..."></textarea>
:::

**c)** Is there statistically significant evidence of a linear association between hours studied and exam score at $\alpha = 0.05$? Justify your answer using the table. (3 pts)

::: short-answer
<textarea placeholder="State decision and reasoning..."></textarea>
:::
:::

::: question
**Q22. (8 pts)** Researchers compare the mean resting heart rate (bpm) of three exercise groups: Sedentary, Moderate, and Active. They use a one-way ANOVA and obtain the following:

| Source | SS | df | MS | F |
|---|---|---|---|---|
| Model | 1240 | 2 | 620 | 8.14 |
| Error | 3050 | 40 | 76.25 | |
| Total | 4290 | 42 | | |

**a)** State $H_0$ and $H_a$ for this test. (2 pts)

::: short-answer
<textarea placeholder="H₀: ...   Hₐ: ..."></textarea>
:::

**b)** What is $R^2$ for this model? Interpret it in context. (3 pts)

::: short-answer
<textarea placeholder="R² = ...  This means ..."></textarea>
:::

**c)** The p-value for F = 8.14 is 0.001. State your conclusion in context at $\alpha = 0.05$, and describe what you would do next. (3 pts)

::: long-answer
<textarea placeholder="We reject/fail to reject H₀ because... Next, we would..."></textarea>
:::
:::

::: question
**Q23. (8 pts)** A researcher wants to test whether the average GPA at a university differs from 3.0. She collects a random sample of $n = 25$ students and finds $\bar{y} = 3.18$ and $s = 0.45$.

**a)** State $H_0$ and $H_a$. (1 pt)

::: short-answer
<textarea placeholder="H₀: μ = ...   Hₐ: μ ≠ ..."></textarea>
:::

**b)** Compute the observed t-statistic. Show your work. (3 pts)

::: short-answer
<textarea placeholder="t = (ȳ - β₀) / (s/√n) = ..."></textarea>
:::

**c)** The p-value is 0.048. At $\alpha = 0.05$, state your decision and conclusion in context. (2 pts)

::: short-answer
<textarea placeholder="Decision: Reject / Fail to reject H₀.  Conclusion: ..."></textarea>
:::

**d)** State one assumption required for the t-distribution to be valid here. (2 pts)

::: short-answer
<textarea placeholder="Assumption: ..."></textarea>
:::
:::

::: question
**Q24. (8 pts)** Explain the difference between a Type I and a Type II error. Then, describe a real-world consequence of each type of error in the context of a clinical trial testing whether a new drug reduces blood pressure. (4 pts each)

::: long-answer
<textarea placeholder="Type I error is...  In this context, it would mean...

Type II error is...  In this context, it would mean..."></textarea>
:::
:::

---

*End of Exam — Total: 130 points possible (100 MC + 30 Short Answer)*

---

```{r answer-key, include=FALSE}
# Answer Key (instructor reference only — set include=TRUE to reveal)
# MC Answers:
# Q1: b   Q2: b   Q3: c   Q4: b
# Q5: b   Q6: b   Q7: c   Q8: b
# Q9: b   Q10: c  Q11: b  Q12: b
# Q13: c  Q14: a  Q15: b  Q16: b
# Q17: a  Q18: b  Q19: b  Q20: b
#
# Short Answer Key Notes:
# Q21a: ŷ = 52.0 + 3.8·(hours)
# Q21b: Each additional hour studied is associated with a 3.8-point increase in exam score.
# Q21c: Yes. t = 3.17, p = 0.005 < 0.05. Reject H₀; significant linear association.
# Q22a: H₀: μ_Sed = μ_Mod = μ_Act; Hₐ: at least one mean differs
# Q22b: R² = 1240/4290 ≈ 0.289; ~28.9% of variability in heart rate explained by exercise group
# Q22c: Reject H₀ (p=0.001 < 0.05). Run pairwise comparisons with corrected α.
# Q23a: H₀: μ = 3.0; Hₐ: μ ≠ 3.0
# Q23b: t = (3.18 - 3.0)/(0.45/√25) = 0.18/0.09 = 2.0
# Q23c: Reject H₀. Significant evidence GPA differs from 3.0 (p=0.048 < 0.05).
# Q23d: Errors are normally distributed OR n is large enough (CLT applies).
# Q24: Type I = false positive (reject true H₀); drug declared effective when it is not.
#       Type II = false negative (fail to reject false H₀); effective drug not approved.
```
