---
title: "Exam 1 Prep"
format: 
  docx:
    reference-doc: refs/custom-reference-doc.docx
  pdf:
    geometry: margin=1in
    keep-tex: false
    toc: false
    number-sections: false
fontsize: 11pt
---

```{r, echo=FALSE}
load('../presentations/data/Fingers.rda')
suppressPackageStartupMessages({
    library(mosaic)
    library(ggformula)
})
```

#### Define **variation** in your own words and explain why it is central to statistics.


#### In the context of a data frame, what do **rows** represent? What do **columns** represent?


#### Which of the following is a *quantitative* variable?

A) Height
B) Favorite color
C) Gender
D) Eye color

#### Why is **sampling** necessary in statistics? Provide one reason.


#### What does it mean for variables to be **categorical** versus **numerical**? Give an example of each from a dataset.

#### In modeling notation, what do the symbols `~` and `data=` indicate? Offer a brief explanation.

#### The vector `x <- c(2,1,3,3,2,3,1,2,1)` is given.

A) After sorting `x`, what pattern becomes visible?
B) What does the frequency table of `x` show?


#### Using the `Fingers` dataset (from class):

A) What do boxplots of `Index ~ Gender` visually display about variability?
B) Describe center, spread, overlap, and any unusual features.

#### Consider the following R code:

```r
gf_histogram(~Score, data = Data, binwidth = 2)
```

A) Interpret what the following R code does and what you would expect the plot to show:

B) What distribution characteristics would the histogram reveal?


#### Explain what the **five-number summary** tells you about a numerical variable and relate it to variation.





#### Write one sentence comparing the distributions displayed by:

```r
gf_histogram(~Index, data = Fingers, binwidth = 0.25)
gf_histogram(~Index, data = Fingers, binwidth = 0.5)
```

Focus on how the choice of **binwidth** affects the appearance.

#### Explain in plain language what an **outlier** is and how the 1.5×IQR rule identifies outliers.


#### What does it tell you if two boxplots (for `Index` by two groups) show a large difference in medians but overlapping boxes? What does that say about variation within and between groups?

#### consider the following R output:

```{r}
favstats(Fingers$Pinkie)
```

A) What can you say about the *center* and *spread* of the `Score` distribution?

B) Use 1.5IQR rule to find if there are any outliers.


#### Explain why it is important to identify the **response variable** before choosing a plot to visualize a relationship.

#### Given these R commands:

```r
mean(Pinkie ~ Gender, data = Fingers) # Returns means for Pinkie in different genders
gf_boxplot(Pinkie ~ Gender, data = Fingers)
```

Explain what each line does and how the two results complement each other.


#### Describe in words what the **residual** represents in the context of a simple model predicting `Pinkie` using `Gender`, and why smaller residuals imply a better model.

#### The following R code calculates proportions:

```{r}
tally(Gender~Job, data=Fingers)
```

Give an example of a conditional probability and compute it.

#### Below is R code that generates a conditional proportion bar chart. Explain how this visualization helps you *explain variability in the response*.

```r
gf_bar(~Job, data = Fingers, fill = ~Gender, position = "fill")
```
```{r}
gf_bar(~Job, data = Fingers, fill = ~Gender, position = "fill")
```

#### Consider the two plots below:

```{r, echo=FALSE}
gf_boxplot(Pinkie ~ Gender, data = Fingers)
gf_bar(~Interest, data = Fingers, fill = ~Gender, position = "fill")

```
Compare and contrast how these two visuals explain variability in the *response* variable. In your answer, mention:

* What the plot shows
* How variation is partitioned or summarized
* What conclusions might be drawn about the role of the explanatory variable

