---
title: "CourseKata Chapter 14"
subtitle: "Targetted Model Comparison"
author: "Mansour Abdoli, PhD"
format:
  live-revealjs:
    slide-number: true
    incremental: true
    chalkboard: true
    preview-links: true
    code-overflow: wrap
    footer: "CourseKata Ch. 14"
engine: knitr
execute:
  echo: true
  warning: false
  message: false
output-files: CK-ch14-brief.html
---

## Overview / Goals

{{< include ./_extensions/r-wasm/live/_knitr.qmd >}}

```{r}
#| echo: false
knitr::opts_chunk$set(echo=FALSE)
suppressPackageStartupMessages({
  library(mosaic)
  library(ggformula)
  library(supernova)
})
def.par <- par(no.readonly = TRUE)
on.exit(par(def.par))

theme_set(
  theme_bw() +
    theme(
    axis.text  = element_text(size = 20),
    axis.title = element_text(size = 22),
    legend.text  = element_text(size = 20),
    legend.title = element_text(size = 22)
  )
)

source('./assets/utils.R')
source('./assets/embedR.R')
source('./data/embedData.R')
load(file = "./data/StudentSurvey.rda")
```

```{r}
#| echo: false
#| results: 'asis'
cat(embedR("./assets/utils.R"))
cat(embedData(StudentSurvey))
```


```{webr}
#| echo: false
#| output: true
#| warning: false
#| message: false

library(mosaic)
library(ggformula)
library(supernova)
options(digits = 3)

theme_set(
  theme_bw() +
    theme(
    axis.text  = element_text(size = 20),
    axis.title = element_text(size = 22),
    legend.text  = element_text(size = 20),
    legend.title = element_text(size = 22)
  )
)

.webr_ready <- TRUE
set.seed(123)
```


:::{.nonincremental}
- Review: Model Overall Significance
  - Simulating $F$ Value
- New Topic: Variable Significance
  - Single-Variable $F$ Distribution
  - Single-Variable $coef$ Distribution
    - Confidence Interval
    - Test of Hypothesis
:::

# Review: Model Overall Significance
## Using `StudentSurvey`
```{webr}
StudentSurvey <- na.omit(StudentSurvey) |> select(Year, Gender, GPA, SAT, HigherSAT)
str(StudentSurvey)
```

## A Complex Model
Let's check significance of $GPA \sim SAT + HigherSAT$

```{webr}
mdl2 <- lm(GPA ~ SAT + HigherSAT, data=StudentSurvey)
summary(mdl2)
```

## Simulating $F$ Value
```{webr}
obs_f <- f(mdl2)
ndo_f <- replicate(100, f(lm(shuffle(GPA) ~ SAT + HigherSAT, data=StudentSurvey)))

gf_histogram(~ ndo_f, bins = 30) |>
  gf_vline(xintercept = obs_f, color = "red")
```

# New Topic: Variable Significance
## Single-Variable $F$ Effect
Base Model (Without $SAT$):
```{webr}
mdl1 <- lm(GPA ~ HigherSAT, data=StudentSurvey)
StudentSurvey$resid1 <- resid(mdl1)
gf_point(resid1~ SAT, data = StudentSurvey, bins = 30) |>
  gf_lm()
```

## Single-Variable $F$ Distribution
Simulation:
```{webr}
obs_SAT_f <- f(lm(resid1 ~ SAT + HigherSAT, data=StudentSurvey))
ndo_SAT_f <- replicate(100, f(lm(shuffle(resid1) ~ SAT + HigherSAT, data=StudentSurvey)))
gf_histogram(~ ndo_SAT_f, bins = 30) |>
  gf_vline(xintercept = obs_SAT_f, color = "red")
```

## Single-Variable $\text{coef}$ Distribution
```{webr}
obs_SAT_bs <- coef(lm(resid1 ~ SAT + HigherSAT, data=StudentSurvey))[['SAT']]
ndo_SAT_bs <- replicate(100, coef(lm(shuffle(resid1) ~ SAT + HigherSAT, data=StudentSurvey))[['SAT']])

gf_histogram(~ndo_SAT_bs, bins=30) |>
  gf_vline(xintercept = obs_SAT_bs, color='red')
```

## Single-Variable $\text{coef}$ C.I.
Finding ME:
```{webr}
mdi <- quantile(ndo_SAT_bs, probs=c(0.025, 0.975))
me <- (mdi[2]-mdi[1])/2 # Margin of Error
cat('ME =', me)
```

95% Confidence Interval:
```{webr}
cat(obs_SAT_bs, ' ± ', me, '\n\n')
cat('(', obs_SAT_bs-me, ', ', obs_SAT_bs+me, ')\n')
```

## Single-Variable $\text{coef}$ ToH
$$p\text{-value}=P(|b_i|>|b_{obs}|)$$
```{webr}
p_val = mean(abs(ndo_SAT_bs)>abs(obs_SAT_bs))
cat('p-Value = ', p_val)
```
