---
title: "CourseKata Chapter 12"
subtitle: "Confidence Interval (1)"
author: "Mansour Abdoli, PhD"
format:
  live-revealjs:
    slide-number: true
    incremental: true
    chalkboard: true
    preview-links: true
    code-overflow: wrap
    footer: "CourseKata Ch. 12 (1)"
engine: knitr
execute:
  echo: true
  warning: false
  message: false
output-files: CK-ch12-sec1-8.html
---

## Overview / Goals

{{< include ./_extensions/r-wasm/live/_knitr.qmd >}}

```{r}
#| echo: false
knitr::opts_chunk$set(echo=FALSE)
suppressPackageStartupMessages({
  library(mosaic)
  library(ggformula)
  library(supernova)
})
def.par <- par(no.readonly = TRUE)
on.exit(par(def.par))

theme_set(
  theme_bw() +
    theme(
    axis.text  = element_text(size = 20),
    axis.title = element_text(size = 22),
    legend.text  = element_text(size = 20),
    legend.title = element_text(size = 22)
  )
)

source('./assets/utils.R')
source('./assets/embedR.R')
source('./data/embedData.R')
load(file = "./data/TipExperiment.rda")
```

```{r}
#| echo: false
#| results: 'asis'
cat(embedR("./assets/utils.R"))
cat(embedData(TipExperiment))
```


```{webr}
#| echo: false
#| output: true
#| warning: false
#| message: false

library(mosaic)
library(ggformula)
library(supernova)
options(digits = 3)

theme_set(
  theme_bw() +
    theme(
    axis.text  = element_text(size = 20),
    axis.title = element_text(size = 22),
    legend.text  = element_text(size = 20),
    legend.title = element_text(size = 22)
  )
)

.webr_ready <- TRUE
set.seed(123)
```


:::{.nonincremental}
- **Sampling Dist. & Test of Hypothesis**
- **Plausible Parameters Values**
- **Confidence Interval**
- **CI & t-Distribution**
- **Assumptions and Interpretation**
:::


# Sampling Dist. & Test of Hypothesis

## Sampling & Null Distributions
- Fixed **Population** $\to$ Fixed **Parameter**
- Random **Sample** $\to$ Random **Statistic**
  - **Sampling Distribution**: </br> The distribution of the random statistic
  - **Null Distribution**: </br> The sampling distribution for a given parameter

## Sampling & Null Distributions
:::{.nonincremental}
:::{.columns}
:::{.column}
- DGP with $\beta_0=\mu=0$

```{webr}
#| fig-width: 5
#| fig-height: 2
#| warning: false
S<- c(-1, 1)
n<- 10 # try different values
gf_histogram(~mean, data=do(1000) * 
  mean(sample(S, n, TRUE)), bins=15) + xlim(-1,1) # coord_cartesian(xlim = c(-1, 1))
```

:::
:::{.column}
- DGP with $\beta_0=\mu=.5$

```{webr}
#| fig-width: 5
#| fig-height: 2
#| warning: false
S<- c(-1, 1, 1, 1)
# reuse `n`
gf_histogram(~mean, data=do(1000) * 
  mean(sample(S, n, TRUE)), bins=15) + xlim(-1,1) # coord_cartesian(xlim = c(-1, 1))
```

:::
:::
:::

## Estimating Null Dist.
Given Sample $\{0, 1, 5, 1, 3\}$

:::{.nonincremental}
:::{.columns}
:::{.column}
- DGP with Sample Mean

```{webr}
#| fig-width: 5
#| fig-height: 2
#| warning: false
s<- c(0, 1, 5, 1, 3); ybar<- mean(s)
n<- length(s)
cat('Observed Mean: ', mean(s), '\n')
gf_histogram(~mean, data=do(1000) * 
  mean(resample(s)), bins=15) + xlim(-1,5)
```

:::
:::{.column}
- DGP with Claimed Mean

```{webr} 
#| fig-width: 5
#| fig-height: 2
#| warning: false
beta0<- 0 # Claimed Mean
s0<- s-ybar+beta0
cat('Claimed Mean: ', mean(s0), '\n')
gf_histogram(~mean, data=do(1000) * 
  mean(resample(s0)), bins=15) + xlim(-3,3)
```

:::
:::
:::


## Plausible Parameters Values
:::{.nonincremental}
:::{.columns}
:::{.column width="55%"}
- Is $\mu=0$ **plausible**?
  - Run a test on $H_0: \mu=0$: </br> $\mu=0$ is *plausible* if we fail to reject.
- process:
  - Define a Sataitic
  - Find **Null Distribution**
  - Reject $H_0$ for extreme observed statistics.

:::
:::{.column width="45%"}
:::{.fragment}
```{webr}
#| fig-width: 4
#| fig-height: 2
#| warning: false
beta0<- 0 # Claimed Mean
cat('Claimed Mean: ', mean(s0), '\n')
dset<- do(1000) * mean(resample(s0))
gf_histogram(~mean, data=dset, bins=15, fill=~(mean>=ybar), show.legend=FALSE) |>
  gf_vline(xintercept = c(0, ybar), color=c('blue', 'red')) + xlim(-3,3)
cat('p-value: ',  mean(dset$mean>=ybar))
```

:::
:::
:::
:::

# Confidence Interval
## Interval Estimate vs. Point Estimate
- $\bar y$ is a **point estimate** (Best Guess) for $\mu$
- An **interval estimate** is all plausible values at $\alpha$.

:::{.fragment}
```{webr}
#| fig-height: 2
#| warning: false
dsetObs<- do(1000) * mean(resample(s))
mus<- c(0.6, 3.6)
dset<- rbind(
  cbind(mu=mus[1], dsetObs-mean(dsetObs$mean)+mus[1]), 
  cbind(mu=mus[2], dsetObs-mean(dsetObs$mean)+mus[2]))
gf_histogram(~mean, data=dset, 
  fill= ~ abs(mu-mean)>=abs(ybar-mu), show.legend=FALSE) |>
  gf_facet_grid(mu~.)|>
  gf_vline(xintercept=ybar, color='red', linetype=2)

dset |> group_by(mu) |> 
  summarize(p.value=mean(abs(mu-mean)>=abs(ybar-mu)))
```
:::

## Distanced Based C.I.
- For symmetric sampling distributions:
$$ C.I. = \text{Point Estimate} \pm \text{Margin of Error}$$

:::{.fragment}
```{webr}
#| fig-height: 2.3
#| warning: false
dsetObs<- do(1000) * mean(resample(s))
quants<- quantile(dsetObs$mean, c(0.025, 0.975))
ME<- diff(quants)/2 # half of middle 95%
cat('ME = ', ME, '\n')
gf_histogram(~mean, data=dsetObs, bins=10,
  fill= ~ (mean<quants[1] | mean>quants[2]), show.legend=FALSE) |>
  gf_vline(xintercept=c(quants, ybar), color=c('blue', 'blue', 'red'), linetype=2) |>
  gf_segment(x = ybar - ME, xend = ybar + ME, y = -2, yend = -2,
  arrow = grid::arrow(end="both", length = unit(.5, "cm")),
  color = "black", linewidth = 1.2) 

```
:::


# CI & t-Distribution
## Theoretical Sampling Distribution
:::{.columns style="font-size:.8em;"}
:::{.column}
- When
  - $Y\sim N(\mu,\sigma)$, or
  - $n$ is large
- $t=\frac{\bar Y-\mu}{SE}\sim t(\text{df}=n-1)$
  - $SE=\frac{s}{\sqrt{n}}$

- For $100(1-\alpha)$% Confidence Level:
  $$ME=t^* \cdot SE$$
  $t^*=t_{\alpha/2}$: top $\alpha/2$ cut-off point
:::
:::{.column}
:::{.fragment}
```{webr}
#| fig-width: 5
#| fig-height: 2
#| warning: false
favstats(s)
SE<- 2/sqrt(5)
t.025<- qt(1-.025, df=5-1)
cat('t*=', t.025, '\n')
ME<- t.025*SE
cint<-c(ybar-ME, ybar+ME)
# cint

```
:::
:::
:::


# Assumptions and Interpretation

## Assumptions

:::{.columns}
:::{.column}

- Independent Observations
- For Theoretical C.I.

:::
:::{.column}
:::
:::

:::{.columns style="font-size:.8em;"}
:::{.column}

  - Is $Y$ normal?  

:::{.fragment}
```{webr, Histogram}
#| fig-width: 5
#| fig-height: 2.5
#| warning: false
gf_dhistogram(~y, data=data.frame(y=s), binwidth=1) |>
  gf_function(fun=dnorm, args=list(mean=mean(s), sd=sd(s)), 
  color='red')

```

```{webr}
#| fig-width: 5
#| fig-height: 2.5
#| warning: false
gf_qq(~s) |> gf_qqline()
```
:::
:::
:::{.column}

  - Is $n$ large?
    - $n>=30$ without outliers
    - Check with/use bootstrap

:::{.fragment}

```{webr}
#| fig-width: 5
#| fig-height: 2.5
#| warning: false
dsetObs<- do(1000) * mean(resample(s))
gf_dhistogram(~mean, data=dsetObs, binwidth=1) |>
  gf_function(fun=dnorm, args=list(mean=mean(s), sd=sd(s)/sqrt(5)), 
  color='red')
```

:::
:::
:::

## Interpretation

- We are $100(1-\alpha)$% confident that the mean is in $(\bar y - ME, \bar y + ME)$.

:::{.fragment}

```{webr}
#| fig-width: 5
#| fig-height: 2.5
#| warning: false
s<- c(0, 1, 5, 1, 3)
ssd<- sd(s); ybar<- mean(s); n<- length(s)

gf_dhistogram(~s, binwidth=1) + xlim(-1, 5)

t.025<- qt(1-.025, df=n-1)
SE<- ssd/sqrt(n); ME<- t.025*SE
gf_histogram(~mean, data=dsetObs, binwidth=1) |>
  gf_function(fun=\(x, ...) 1000*dnorm(x,...), args=list(mean=mean(s), sd=sd(s)/sqrt(5)), 
  color='red') |>
  gf_segment(x = ybar - ME, xend = ybar + ME, y = -2, yend = -2,
  arrow = grid::arrow(end="both", length = unit(.5, "cm")),
  color = "black", linewidth = 1.2)  + xlim(-1, 5)

quantile(s, c(.025, 0.975))
quantile(dsetObs$mean, c(.025, 0.975))

```

:::

## Caution
- C.I is for the population mean, not observations.
- C.I is a random variable, the populaiton mean is fixed.

:::{.fragment}

```{r}
#| fig-width: 5
#| fig-height: 2.5
#| warning: false
s<- TipExperiment$Tip
ssd<- sd(s); ybar<- mean(s); n<- length(s)
xlim<- range(s)

gf_dhistogram(~s, binwidth=1) + xlim(xlim)

t.025<- qt(1-.025, df=n-1)
SE<- ssd/sqrt(n); ME<- t.025*SE

dsetObs<- do(1000) * mean(resample(s))
gf_histogram(~mean, data=dsetObs, binwidth=1) |>
  gf_function(fun=\(x, ...) 1000*dnorm(x,...), args=list(mean=mean(s), sd=sd(s)/sqrt(5)), 
  color='red') |>
  gf_segment(x = ybar - ME, xend = ybar + ME, y = -2, yend = -2,
  arrow = grid::arrow(end="both", length = unit(.5, "cm")),
  color = "black", linewidth = 1.2)  + xlim(xlim)

quantile(s, c(.025, 0.975))
quantile(dsetObs$mean, c(.025, 0.975))

```

:::
