---
title: "CourseKata Chapter 13"
subtitle: "Multivariate Models"
author: "Mansour Abdoli, PhD"
format:
  live-revealjs:
    slide-number: true
    incremental: true
    chalkboard: true
    preview-links: true
    code-overflow: wrap
    footer: "CourseKata Ch. 13 (1)"
engine: knitr
execute:
  echo: true
  warning: false
  message: false
output-files: CK-ch13-sec1-6.html
---

## Overview / Goals

{{< include ./_extensions/r-wasm/live/_knitr.qmd >}}

```{r}
#| echo: false
knitr::opts_chunk$set(echo=FALSE)
suppressPackageStartupMessages({
  library(mosaic)
  library(ggformula)
  library(supernova)
})
def.par <- par(no.readonly = TRUE)
on.exit(par(def.par))

theme_set(
  theme_bw() +
    theme(
    axis.text  = element_text(size = 20),
    axis.title = element_text(size = 22),
    legend.text  = element_text(size = 20),
    legend.title = element_text(size = 22)
  )
)

source('./assets/utils.R')
source('./assets/embedR.R')
source('./data/embedData.R')
load(file = "./data/Smallville.rda")
```

```{r}
#| echo: false
#| results: 'asis'
cat(embedR("./assets/utils.R"))
cat(embedData(Smallville))
```


```{webr}
#| echo: false
#| output: true
#| warning: false
#| message: false

library(mosaic)
library(ggformula)
library(supernova)
options(digits = 3)

theme_set(
  theme_bw() +
    theme(
    axis.text  = element_text(size = 20),
    axis.title = element_text(size = 22),
    legend.text  = element_text(size = 20),
    legend.title = element_text(size = 22)
  )
)

.webr_ready <- TRUE
set.seed(123)
```


:::{.nonincremental}
- Does more Predictor mean better Model?
- Explaning Variability by Different Predictors
- Visualizing Variability by Multiple Predictors
- Constructing a Multivariate Model
- Interpreting Coefficients
- Comparing Models
:::

# Does more Predictor mean better Model?

## Model Complexity Review
$$
\text{DATA} = \text{MODEL} + \text{ERROR}
$$

- Empty Model: Mean Response
- Simple Model: 2-Group Means or a Simple Regression
- Complex: Multiple-Group Means and/or more Numerical variables

## The Smallville data

Outcome:

:::{style="font-size:0.85em;"}

- `PriceK`: sale price in thousands of dollars
:::

Predictors:

:::{style="font-size:0.85em;"}

- `Neighborhood`: Downtown or Eastside
- `HomeSizeK`: square footage in thousands
- `HasFireplace`: fireplace indicator

:::


```{webr}
#| label: inspect-smallville
head(Smallville)
```


# Explaning Variability by Different Predictors

## Variability in `PriceK`

::: {.columns}
::: {.column width="33%"}
```{webr}
#| fig-width: 3.5
#| fig-height: 3.5
null_model <- lm(PriceK~NULL, data=Smallville)
gf_jitter(PriceK~'Overall', data=Smallville, width=.2) |> gf_labs(x='') |> 
gf_model(null_model)
```
:::
::: {.column width="33%"}
```{webr}
#| fig-width: 3.5
#| fig-height: 3.5
Neighborhood_model <- lm(PriceK ~ Neighborhood, data = Smallville)
gf_jitter(PriceK~Neighborhood, data=Smallville, width=.15, color=~Neighborhood, show.legend=FALSE) |> 
  gf_model(Neighborhood_model)
```
:::
::: {.column width="33%"}
```{webr}
#| fig-width: 3.5
#| fig-height: 3.5
HomeSizeK_model <- lm(PriceK ~ HomeSizeK, data = Smallville)
gf_point(PriceK~HomeSizeK, data=Smallville) |> 
  gf_model(HomeSizeK_model)
```
:::
:::

## Two single-predictor models

:::{style="font-size:0.9em;"}
Which model reduces more error; explains more variability?
:::

:::{.fragment style="font-size=0.85em;"}
```{webr}
#| label: single-predictor-models

supernova(Neighborhood_model)
cat('\n==========\n')
supernova(HomeSizeK_model)
```
:::


# Visualizing Variability by Multiple Predictors

## Using Both Predictors
::: {.columns style="font-size:0.85em;"}
::: {.column}
- Does neighborhood add any value?
:::
::: {.column}
- Does home-size effect changes?
:::
:::

::: {.columns}
::: {.column}
:::{.fragment}
```{webr}
#| label: color-neighborhood
#| fig-width: 5
#| fig-height: 3.5
gf_point(PriceK ~ HomeSizeK, data = Smallville,
         color = ~ Neighborhood, show.legend=FALSE) %>%
  gf_model(HomeSizeK_model, color = "black")
```
:::


:::
::: {.column}
:::{.fragment}
```{webr}
#| label: facet-neighborhood
#| fig-width: 5
#| fig-height: 3.5
gf_point(PriceK ~ HomeSizeK, data = Smallville) %>%
  gf_facet_grid(. ~ Neighborhood)
```

:::
:::
:::

# Constructing a Multivariate Model

## Specifying a multivariate model

$$
Y_i = b_0 + b_1X_{1i} + b_2X_{2i} + e_i
$$

For Smallville:

$$
\text{PriceK}_i = b_0 + b_1\text{Neighborhood}_i + b_2\text{HomeSizeK}_i + e_i
$$

## Fitting the model in R

```{webr}
#| label: fit-multivariate
multi_model <- lm(PriceK ~ Neighborhood + HomeSizeK, data = Smallville)
multi_model
```


- Model Fit:  Minimize **SS Error**

## Write the fitted model

```{webr}
#| label: coefficients
coef(multi_model)
```

$$
\widehat{\text{PriceK}} = b_0 + b_1(\text{NeighborhoodEastside}) + b_2(\text{HomeSizeK})
$$

:::{.callout-important}
`NeighborhoodEastside = 1` for Eastside and `0` for Downtown.
:::



# Interpreting Coefficients

## Interpreting the intercept

- The intercept is the predicted price when:

  - `NeighborhoodEastside = 0` → Downtown
  - `HomeSizeK = 0`

- Intercept is the model's Downtown baseline at 0K square feet.

- *Is this substantively meaningful, or mainly mathematical?*

## Interpreting `NeighborhoodEastside`

- The *predicted* difference in price ...
  - between Eastside and Downtown homes ...
  - **holding home size constant** ...
  - in $1000\$$ is ...

:::{.fragment}
```{webr}
#| label: neighborhood-coef
coef(multi_model)["NeighborhoodEastside"]
```
:::

## Holding the other variable constant

This means we compare observations that are the same on the other predictor.

- Same home size, different neighborhood
- Same neighborhood, different home size

# Comparing Models

## Compare single vs multivariate slope

:::{.fragment}

Single Predictor Model:

```{webr}
#| label: compare-slopes0
coef(HomeSizeK_model)
```

Two Predictor Model:

```{webr}
#| label: compare-slopes1
coef(multi_model)
```

:::

- Why does the `HomeSizeK` coefficient get smaller after adding `Neighborhood`?


## Predictions & Residuals

```{webr}
#| label: predictions
Smallville$multi_predict <- predict(multi_model)
Smallville$multi_resid <- resid(multi_model)

Smallville %>%
  select(PriceK, Neighborhood, HomeSizeK, multi_predict, multi_resid) %>%
  head()
```

## Visualizing predictions

```{webr}
#| label: prediction-triangles
#| fig-width: 7
#| fig-height: 4
gf_point(PriceK ~ HomeSizeK, color = ~ Neighborhood, data = Smallville) %>%
  gf_point(multi_predict ~ HomeSizeK, data = Smallville,
           color = "black", shape = 2) |>
  gf_lm(multi_predict ~ HomeSizeK, color = ~ Neighborhood) |> #gf_model(multi_model) |>
  gf_labs(subtitle='The triangles are the model predictions.')
```


## Parallel lines

:::{.columns}
:::{.column}
:::{.fragment}

```{webr}
#| label: parallel-lines
#| fig-width: 5
#| fig-height: 4
gf_point(PriceK ~ HomeSizeK, color = ~ Neighborhood, data = Smallville, show.legend=FALSE) %>%
  gf_lm(multi_predict ~ HomeSizeK, color = ~ Neighborhood, show.legend=FALSE) #gf_model(multi_model)
```

:::
:::
:::{.column}

- For **additive models**, group lines have:
  - the **same slope** and
  - different **intercepts**.
:::
:::

## Residuals still mean the same thing

$$
\text{Residual} = \text{Observed} - \text{Predicted}
$$

```{webr}
#| label: residual-example
Smallville %>%
  mutate(check = multi_predict + multi_resid) %>%
  select(PriceK, multi_predict, multi_resid, check) %>%
  head()
```


## Sum of squared residuals

SS Error: Error left after using both predictors.
```{webr}
#| label: ss-error from residual
sum(resid(multi_model)^2)
```

```{webr}
#| label: ss-error from ANOVA
supernova(multi_model)
```



## Comparing models by error reduction

:::{style="font-size:.8em;"}
Which model is **better**?

Could the difference be by **chance**?


```{webr}
#| label: Neighborhood only
supernova(Neighborhood_model)
```
```{webr}
#| label: Home Size only
supernova(HomeSizeK_model)
```
```{webr}
#| label: Both Predictors
supernova(multi_model)
```

:::
