---
title: "CourseKata Chapter 9 (1 to 8)"
subtitle: "Linear Regression"
author: "Mansour Abdoli, PhD"
format:
  live-revealjs:
    slide-number: true
    incremental: true
    chalkboard: true
    preview-links: true
    code-overflow: wrap
    footer: "CourseKata Ch. 9"
engine: knitr
execute:
  echo: true
  warning: false
  message: false
output-files: CK-ch9-sec1-8.html
---

## Overview / Goals

{{< include ./_extensions/r-wasm/live/_knitr.qmd >}}

```{r}
#| echo: false
knitr::opts_chunk$set(echo=FALSE)
suppressPackageStartupMessages({
  library(mosaic)
  library(ggformula)
  library(supernova)
})
def.par <- par(no.readonly = TRUE)
on.exit(par(def.par))

source('./assets/utils.R')
source('./assets/embedR.R')
source('./data/embedData.R')
load(file = "./data/Fingers.rda")
```

```{r}
#| echo: false
#| results: 'asis'
cat(embedR("./assets/utils.R"))
cat(embedData(Fingers))
```


```{webr}
#| echo: false
#| output: true
#| warning: false
#| message: false

library(mosaic)
library(ggformula)
library(supernova)
options(digits = 3)

.webr_ready <- TRUE
set.seed(123)
```


:::{.nonincremental}
By the end of today, you can:

- fit and interpret **multi‑group (3+ groups)** models using `lm()`
- compare group models using **PRE**, **df**, **MS**, and the **F ratio**
- describe and compute **effect sizes** (mean differences, PRE, Cohen’s *d*)
- use **shuffle/simulation** to ask whether an effect could plausibly be due to randomness

:::

## Review: Group Models
:::{.nonincremental}
- Data:
  - Outcome: $Y$ (A numerical Variable)
  - Explanatory: A Categorical Variable
- Models:
  - **Empty model**: one mean for everyone
  - **Group model**: *separate mean* for each group.

:::

## Example

::: {.columns}
::: {.column width='33%'}
```{webr}
#| fig-height: 3
#| fig-width: 3
gf_jitter(Thumb ~ 'NULL', data = Fingers, width = 0.15, alpha = 0.6) |> 
gf_model(lm(Thumb~NULL, data=Fingers), color='orange')
```
:::
::: {.column width='33%'}
```{webr}
#| fig-height: 3
#| fig-width: 3
gf_jitter(Thumb ~ Gender, data = Fingers, width = 0.15, alpha = 0.6) |> 
gf_model(lm(Thumb~Gender, data=Fingers), color='orange')
```
:::
::: {.column width='33%'}
```{webr}
#| fig-height: 3
#| fig-width: 3 
gf_jitter(Thumb ~ Interest, data = Fingers, width = 0.15, alpha = 0.6) |> 
gf_model(lm(Thumb~Interest, data=Fingers), color='orange')
```
:::
:::
<!-- end columns -->


## Comparing Models

::: {.columns}
::: {.column}
:::{style="font-size: .5em;"}

```{webr}
cat('Gender: 2-Level Category Explanatory:')
supernova(lm(Thumb~Gender, data=Fingers))
cat('\nInterest 3-Level Category Explanatory:')
supernova(lm(Thumb~Interest, data=Fingers))
```

:::
:::
::: {.column}

```{r, echo=FALSE}
modG <- lm(Thumb~Gender, data=Fingers)
modI <- lm(Thumb~Interest, data=Fingers)
```

:::{.fragment .nonincremental}

- Model Measures:

```{r}
data.frame(
  Gender = c(sse(modG), pre(modG), f(modG)) |>
    formatC(digits = 5),
  Interest = c(sse(modI), pre(modI), f(modI)) |>
    formatC(digits = 5),
  row.names = c('SSE', 'PRE', 'F')
)

```

- Which one is better?
:::
:::
:::
<!-- end columns -->

# Effect of Increasing Levels
## Example: 2-Level Explanatory

::: {.columns}
::: {.column}
```{webr}
#| fig-height: 3
#| fig-width: 5
Fingers$Height2 <- ntile(Fingers$Height, n=2) |>
  factor(1:2, c('short', 'tall'))
mod2 <- lm(Thumb~Height2, data=Fingers)
gf_jitter(Thumb ~ Height2, data = Fingers, width = 0.15, alpha = 0.6) |> 
  gf_model(mod2, color='orange')
```
:::
::: {.column style="font-size: .7em;"}
::: {.fragment .nonincremental style="font-size: .7em;"}
- $Y_i = b_0 + b_1 \cdot X_{1i} + e_i$ </br>
  - $b_0 = \text{mean}(\text{Thumb}_\text{short})$
  - $b_1 = \text{mean}(\text{Thumb}_\text{tall})-\text{mean}(\text{Thumb}_\text{short})$
  - $X_{1i} = 1 \text{ for Height}==\text{tall}$ 
:::
:::{.fragment}
```{webr}
mod2
```
:::
:::
:::

## Example: 3-Level Explanatory

::: {.columns}
::: {.column}
```{webr}
#| fig-height: 3
#| fig-width: 5
Fingers$Height3 <- ntile(Fingers$Height, n=3) |>
  factor(1:3, c('Short', 'Med.', 'Tall'))
mod3 <- lm(Thumb~Height3, data=Fingers)
gf_jitter(Thumb ~ Height3, data = Fingers, width = 0.15, alpha = 0.6) |> 
  gf_model(mod3, color='orange')
```
:::
::: {.column style="font-size: .7em;"}
::: {.fragment .nonincremental style="font-size: .7em;"}
- $Y_i = b_0 + b_1 \cdot X_{1i} + b_2 \cdot X_{2i} + e_i$ </br>
  - $b_0 = \text{mean}(\text{Thumb}_\text{Short})$
  - $b_1 = \text{mean}(\text{Thumb}_\text{Med.})-\text{mean}(\text{Thumb}_\text{Short})$
    - $X_{1i} = 1 \text{ for Height}==\text{Med.}$ 
  - $b_2 = \text{mean}(\text{Thumb}_\text{Tall})-\text{mean}(\text{Thumb}_\text{Short})$
    - $X_{2i} = 1 \text{ for Height}==\text{Tall}$ 
:::

:::{.fragment}
```{webr}
mod3
```
:::
:::
:::

<!-- end columns -->

## Comparing Model Measures
::: {.columns style="font-size:.7em"}
::: {.column width="65%"}
```{webr}
supernova(mod2)

cat('\n')
supernova(mod3)
```
:::
::: {.column width="35%"}
- Notice anything?
:::
:::
<!-- end columns -->

## Level Size and Model Measures

::: {.columns style="font-size: .9em;"}
::: {.column}
### SSE (SSR):
- How much of error is not explained.

:::
::: {.column}

### PRE = (1-SSE)/SST
- Proportion of Explained variability

:::
:::

  - More levels, more complex models, more explained variability

::: {.fragment style="font-size: .9em;"}
### F = MSM/MSE
:::

:::{style="font-size: .9em;"}
- Ratio of average explained variability to unexplained variablity
  - Less sensetive to larger number of levels
:::

# The F Ratio

## Why we need F

:::{style="font-size: .85em;"}
PRE alone doesn’t “penalize” complexity.

- Adding parameters almost always reduces SSE.
  - *Is the reduction big enough to be worth the extra parameters?*

:::{.fragment}
### F ratio $$F = \frac{MS_{\text{Model}}}{MS_{\text{Error}}}$$
:::

- Large $F$ means: *error reduced per parameter is large relative to remaining error.*
:::

# Effect Size in Group Models

## Why effect size?

- With large $n$, tiny effects can look convincing.
- Effect size says: *How big is the difference?*

::: {.fragment}
Common group-model effect sizes:
:::

- **Mean differences** (model parameters)
- **PRE** (variance explained relative to empty model)
- **Cohen’s $d$** (standardized mean difference)

## Cohen’s d (two groups)
:::{.columns}
:::{.column width="35%"}
:::{style="font-size:.7em;"}
- Standardized Difference: $$d = \frac{\bar Y_1 - \bar Y_2}{s_{pooled}}$$
  - $s_{pooled}^2 = MSE = \frac{SSE}{df_E}$
  - In general: $s_{pooled}^2 = \frac{df_1\cdot s_1^2+df_2\cdot s_2^2}{df_1+df_2}$
:::
:::
:::{.column width="65%"}
:::{.fragment style="font-size:.7em;"} 
```{webr}
modG <- lm(Thumb~Gender, data=Fingers)
coef(modG)
supernova(modG)
```
:::
:::
:::

:::{.columns}
:::{.column width="35%"}
:::{.fragment style="font-size:.8em;"}
```{webr}
cohensD(modG)
```
:::
:::

:::{.column width="65%"}
:::{style="font-size:.8em;"}
  - Verify the calculation.
:::
:::
:::

## Effect size ≠ causality

- Large difference between groups:
  - Practical Effect
  - Nor a proof of causal effect

\

- Group models show association
- Causal claims need stronger design / evidence

# Thinking about DGP

## Observed Sample to Population Inference

- The observed effect:
  - A real systematic difference,
  - Or a **random variation** (noise)

- A Test Approach
  - Assume *random variation* (the true effect = 0) 
    - Null Hypothesis ($H_0$)
  - Find Null Distribution of effect
    - Sampling Distribution under the Null
  - Measure how often effects are as large as observed one

# Shuffle-based Simulation (Conceptual Inference)

## Shuffle logic

- Keep the same $Y$ values
- Randomly reassign them to groups
- Refit the model
- Record the effect size (e.g., $b_1$ or PRE)
- Repeat many times to get a **shuffle distribution**
  - A.k.a. **Null Distribution**

## Example: Two-group Model

:::{.fragment}
```{webr}
# To get the same results
set.seed(123) 
# observed effect
m_obs <- lm(Thumb ~ Sex, data = Fingers)
b_obs <- b1(m_obs)

# shuffle distribution of b1
shuf_b1 <- rep(NA, 500)
for(k in 1:500){
  d <- Fingers
  d$Thumb <- shuffle(d$Thumb)
  shuf_b1[k] <- b1(lm(Thumb ~ Sex, data = d))
}

c(observed_b1 = b_obs,
  shuf_mean = mean(shuf_b1),
  shuf_sd = sd(shuf_b1))
```
:::

## Visualize the shuffle distribution

:::{.fragment}
```{webr}
#| fig-height: 3
#| fig-width: 7

gf_histogram(~shuf_b1, bins = 30) %>%
  gf_vline(xintercept = b_obs)
```
:::

## Interpreting the shuffle plot

:::{.nonincremental}
- The histogram:
  - **effects expected by chance** <br/>(if the DGP had no group effect)
- The vertical line is the **observed effect**
:::

:::{.fragment .nonincremental}
- If the observed effect is far out in the tails:  
  - The “no-effect” DGP is **unlikely** 
:::

# Wrap-up

## What Chapter 8 emphasizes
:::{style="font-size:0.85em;"}
- **Multi-group models**: <br/> baseline mean and group changes
- **Model comparison**: <br/>fit (PRE) + complexity (df)
- **F ratio**: <br/>a complexity-aware comparison tool
- **Effect sizes**: <br/> Statistical vs Practical Difference
- **Shuffle simulation**: <br/>Could the effect be due to Noise?
:::
