---
title: "CourseKata Chapter 11"
subtitle: "Model Significance (2)"
author: "Mansour Abdoli, PhD"
format:
  live-revealjs:
    slide-number: true
    incremental: true
    chalkboard: true
    preview-links: true
    code-overflow: wrap
    footer: "CourseKata Ch. 11 (2)"
engine: knitr
execute:
  echo: true
  warning: false
  message: false
output-files: CK-ch11-sec11-13.html
---

## Overview / Goals

{{< include ./_extensions/r-wasm/live/_knitr.qmd >}}

```{r}
#| echo: false
knitr::opts_chunk$set(echo=FALSE)
suppressPackageStartupMessages({
  library(mosaic)
  library(ggformula)
  library(supernova)
})
def.par <- par(no.readonly = TRUE)
on.exit(par(def.par))

theme_set(
  theme_bw() +
    theme(
    axis.text  = element_text(size = 20),
    axis.title = element_text(size = 22),
    legend.text  = element_text(size = 20),
    legend.title = element_text(size = 22)
  )
)

source('./assets/utils.R')
source('./assets/embedR.R')
source('./data/embedData.R')
load(file = "./data/Fingers.rda")
```

```{r}
#| echo: false
#| results: 'asis'
cat(embedR("./assets/utils.R"))
cat(embedData(Fingers))
```


```{webr}
#| echo: false
#| output: true
#| warning: false
#| message: false

library(mosaic)
library(ggformula)
library(supernova)
options(digits = 3)

theme_set(
  theme_bw() +
    theme(
    axis.text  = element_text(size = 20),
    axis.title = element_text(size = 22),
    legend.text  = element_text(size = 20),
    legend.title = element_text(size = 22)
  )
)

.webr_ready <- TRUE
set.seed(123)
```


:::{.nonincremental}
- **Complex v. Simple Models**
- **Type I Error Inflation**
- **Single Model Test**
- **Pairwise Comparisons**
- **Test of Independence**
:::


## Test Procedure Workflow

:::{.nonincremental}
1. State the null & Alt. hypothesis
2. Find the null distribution of the test statistic
3. Compute the observed (test) statistic
4. Compute p-value or check against R.R.
5. Make a decision & write a conclusion
6. Check assumptions
:::

# Complex v. Simple Models
## Empty Model vs 2-Parameter Model 
- Empty Model: $$Y = \beta_0 + \varepsilon$$
- Binary Explanatory: $$Y = \beta_0 + \beta_1 X + \varepsilon, \quad \beta_0=\mu_1, \beta_1=\mu_2-\mu_1$$
- Simple Linear Regression: $$Y = \beta_0 + \beta_1 X + \varepsilon, \quad \ \beta_0=\mu_{|X=0},\  \beta_1=\frac{\Delta \mu}{\Delta X}$$

## Complex Models
Models with more than two parameters:

- Multi-level Categorical Explanatory: 
$$\text{Thumb}=\beta_0 + \beta_1 X_{[\text{Part-time Job}]} + \beta_2 X_{[\text{Full-time Job}]} + \varepsilon$$
- Multiple Numerical Explanatory:
$$\text{Thumb}=\beta_0 + \beta_1 \text{Height} + \beta_2 \text{GradePredict} + \varepsilon$$
- Combination of Numerical and Categorical: $$\text{Thumb}\sim \text{Height}+\text{Job}$$

## Testing Complex vs Empty Models

:::{style="font-size:.8 em;"}
- Testing non-empty model parameters one at a time
  $$\begin{align} H_0:& \beta_1=0\text{ vs. }H_a: \beta_1\ne 0\\
  \vdots&\\
  H_0:& \beta_k=0\text{ vs. }H_a: \beta_k\ne 0
  \end{align}$$
- Testing all model at once
  $$\begin{align} H_0:& \beta_1=\cdots=\beta_k=0 \\ H_a:& \text{At least one }\beta_i\ne 0, i=1, \cdots, k\end{align} $$
:::

## Testing Two Complex Models
- Testing Model, so far:
  - Model vs Empty Model
- Other types of comparisons
  - Complex vs Simple Models
  
- Example:
  - $\text{Thumb}\sim \text{Gender}$ vs. $\text{Thumb}\sim \text{Job}$
  - $\text{Thumb}\sim \text{Heigh}$ vs. $\text{Thumb}\sim \text{Heigh} + \text{Gender}$

# Type I Error Inflation
## $k$ Simultaneous Tests
When testing $k$ separate tests at $\alpha$ sig. level:

- For each test: $P(\text{Type I Error})=\alpha$
- The over all Type I Error:
  - To reject at least one in error, which is
  - The complement of rejecting all correctly; that is,

:::{.fragment}
  $$\begin{align} P(\text{Overall Type I Error})=& 1-P(\text{All correctly rejected})\\
  =&1-(1-\alpha)^k\end{align}$$
:::

## Mulriple Tests Correction
- Simultaneous multiple tests **inflates** Type I Error.
- Bonferroni's Correction for $g$ comparisons:
  - For each test:
    - Use $\alpha/g$ as *Significance Level*
    - Report $g\times p\text{-value}$
  - Good for small $g$
- Tukey's Correction
  - More complex and more powerful
  
# Single Model Test
## PRE and $F$ Tests
- PRE and $F$:
  - The whole model performance
  - Testing all parameters at once
  - Comparing complex vs simple models
- Test Procedure
  - Assumption $\to$ Null Distribution $\to$ Observed Value

## $F$ Null Distributions
- Simulation Null Distribution
  - Assumption: No Association
  - Distribution: Histogram of many replications
- Theoretical Null Distribution
  - Assumption: No Association & $\varepsilon\sim N(0, \sigma)$
  - Distribution: $\text{F}\sim \text{F}\left(DF_M, DF_E\right)$
- Used in ANOVA

## $F$ Theoretical Distribution Example
:::{.columns}
:::{.column width="35%"}
```{webr}
#| fig-height: 3
#| fig-width: 3
dfM <- 2 #length(levels(Fingers$Job))-1
dfE <- 154 #dim(Fingers)[1] - 1 - dfM
null_f <- do(500) * f(shuffle(Thumb)~Job, data=Fingers)
gf_dhistogram(~f, data=null_f, bins=30) |>
  gf_function(fun = df, args=list(df1=dfM, df2=dfE), color='red') |>
  gf_point(0~x, data=data.frame(x=f(Thumb~Job, data=Fingers)))
```
:::
:::{.column width="65%"}
```{webr}
#| fig-height: 3
#| fig-width: 7
cat('P(F>obs.F)=', 1-pf(f(Thumb~Job, data=Fingers), dfM, dfE), '\n')
supernova(m1)

```
:::
:::


## PRE Null Distributions
- Simulation Null Distribution
  - Assumption: No Association
  - Distribution: Histogram of many replications
- Theoretical Null Distribution
  - Assumption: No Association & $\varepsilon\sim N(0, \sigma)$
  - Distribution: $\text{PRE}\sim \text{Beta}\left(\alpha=\frac{DF_M}{2}, \beta=\frac{DF_E}{2}\right)$


## PRE Theoretical Distribution Example
:::{.columns}
:::{.column width="35%"}
```{webr}
#| fig-height: 3
#| fig-width: 3
dfM <- 2; dfE <- 154
obs_pre <- pre(Thumb~Job, data=Fingers)
null_pre <- do(500) * pre(shuffle(Thumb)~Job, data=Fingers)
gf_dhistogram(~pre, data=null_pre, bins=30) |>
  gf_function(fun = dbeta, args=list(shape1=dfM/2, shape2=dfE/2), color='red') |>
  gf_point(0~x, data=data.frame(x=obs_pre))
```
:::
:::{.column width="65%"}
```{webr}
#| fig-height: 3
#| fig-width: 7
cat('obs.PRE=', obs_pre, '\n')
cat('P(PRE>obs.PRE)=', 1-pbeta(obs_pre, dfM/2, dfE/2), '\n')

```
:::
:::

## Testing Complex vs Simple
- $F$ and PRE:
  - Use Variations reduction from Empty Model
- Comparing Complex vs. Simple
  - Error reduction $= SSE_{Simple} - SSE_{Complex}$
  - DF of the reduction $= DFE_{Simple} - DFE_{Complex}$
- Compare Complex Model:
  - Use PRE & $F$ of Reduction

## Example: Comparing Complex vs Simple
```{webr}
m1 <- lm(Thumb~Job, data=Fingers)
m2 <- lm(Thumb~Job + Height, data=Fingers)
supernova(m1)
supernova(m2)
anova(m1, m2)
cat('\nP(F>', 26.846, ') =', 1-pf(26.846, 1, 153), '\n')
```

# Pairwise Comparisons
## Controlling Type I Error
- Test the whole model at once ($F$ or $PRE$)
  - If Significant, run **post-hoc** analysis
- Pairwise Post-hoc Tests:
  - Multiple Levels (for categorical)
  - Multiple Variables (for numerical and categorical)

## Example: Post-hoc on Categorical
```{webr}
# Thumb~Job is not significant
# Thumb~RaceEthnic is significant at 7%
m.race<- lm(Thumb~RaceEthnic, data = Fingers)
supernova(m.race)
cat('\nBonferroni Correction:\n')
pairwise(m.race, correction = "Bonferroni")
cat('\nTukey Correction:\n')
pairwise(m.race, correction = "Tukey")
```

# Test of Independence
## Recall: Test Procedure Workflow 

:::{.nonincremental}
1. State the null & Alt. hypothesis
2. Find the null distribution of the test statistic
3. Compute the observed (test) statistic
4. Compute p-value or check against R.R.
5. Make a decision & write a conclusion
6. Check assumptions
:::

> Works for any **test statistics**

## Independence of Categorical Variables
Example: Are `Gender` and `Job` independent?

:::{.columns}
:::{.column}
```{r}
#| fig-width: 5
gf_bar(~ Gender, fill = ~ Job, data = Fingers, position='fill') %>%
  gf_labs(
    title = "Association Between Gender and Job",
    x = "Gender",
    y = "Count",
    fill = "Job"
  )
```
:::
:::{.column}
```{r}
#| fig-width: 5
prop.table(table(Fingers$Job, Fingers$Gender), margin = 2)
```
:::
:::

## What does Independent Means?
:::{.columns}
:::{.column}
Observed Proportions:
```{r}
prop.table(table(Fingers$Job, Fingers$Gender), margin = 2)
```

:::
:::{.column}
Observed Counts $n_{ij}$:
```{r}
#| fig-width: 5
table(Fingers$Job, Fingers$Gender)
```
:::
:::

Expected Counts under the Null (Independence):

| Job |female|male| Total|
|-----------|-----|-----|-----|
|Not Working|  |  |  |
|Part-time Job|  |  |  |
|Full-time Job|  |  |  |
| Total|  |  |  |

## Deviasion-based Test Statistic 
- Expected Counts: $e_{ij}=\frac{n_{i+}\times n_{+j}}{n}$
- Deviation: $n_{ij}-e_{ij}$
- Test Statistic: $\chi^2 = \sum_{all\ cells} \frac{(n_{ij}-e_{ij})^2}{e_{ij}}$
- Assumption: Independence & Large $n_{ij}$:
  -$\chi^2\sim \text{ch-square}\left(df=(n_{row}-1)(n_{col}-1)\right)$

::: {.fragment}
```{webr}
chisq.test(Fingers$Gender, Fingers$Job)
```
:::

## Simulation-Based Test of Independence
- Test Statistic: $\chi^2 = \sum_{all\ cells} \frac{(n_{ij}-e_{ij})^2}{e_{ij}}$
- Assumption: Independence
  - Shuffle one variable
  - Calculate test statistics
  - Repeat many times
  - Create a histogram as Null Distributoin

::: {.fragment}
```{webr}
obs_chi2<- chisq(Job~Gender, data=Fingers)
null_chi2 <- do(500) * chisq(shuffle(Job)~Gender, data=Fingers)
gf_dhistogram(~X.squared, data = null_chi2, bins=30) |>
  gf_function(fun = dchisq, args=list((2-1)*(3-1)), color='red') |> gf_point(0~x, data=data.frame(x=obs_chi2), color = 'red')
```
:::

