---
title: "CourseKata Chapter 7"
subtitle: "Modeling by Group Means"
author: "Mansour Abdoli, PhD"
format:
  live-revealjs:
    slide-number: true
    incremental: true
    chalkboard: true
    preview-links: true
    code-overflow: wrap
    footer: "CourseKata Ch. X"
#resources:
#  - assets/webr_helpers.R
engine: knitr
execute:
  echo: true
  warning: false
  message: false
output-files: CK-ch7-complete.html
---

## Session Goals

{{< include ./_extensions/r-wasm/live/_knitr.qmd >}}

```{r}
#| echo: false
suppressPackageStartupMessages({
  library(mosaic)
  library(ggformula)
})
```

```{r}
#| echo: false
#| results: 'asis'
# create a webr to read the data
source('./assets/utils.R')
source('./assets/embedR.R')
source('./data/embedData.R')
# for more data:
## load(file =) # load a dataset
## cat(embedData(dataset)) # Embed it 
load(file = "./data/Fingers.rda")
```

```{r}
#| echo: false
#| results: 'asis'
cat(embedR("./assets/utils.R"))
cat(embedData(Fingers))
```


```{webr}
#| echo: false
#| output: false
#| warning: false
#| message: false

library(mosaic)
library(ggformula)
options(digits = 3)

# Load helper functions if present (optional)
helpers <- readLines("assets/webr_helpers.R", warn = FALSE)
eval(parse(text = helpers))

Fingers$MathAnxious <- 
  factor(Fingers$MathAnxious,
  levels=c('Strongly Disagree', 
    'Disagree', 
    'Neither Agree nor Disagree', 
    'Agree', 'Strongly Agree'))
Fingers$Job <- 
  factor(Fingers$Job,
         c("Not Working", "Part-time Job", "Full-time Job"))

.webr_ready <- TRUE
```

:::{.nonincremental}
By the end of today you can:

* Explain what a group model does
* Fit `lm(y ~ group)`
* Interpret intercept and coefficients
* Explain how predictions change from the empty model

:::

## Review: The Empty Model

- DATA = Mean + Error

:::{.fragment .nonincremental}
- Assumption:
  - Everyone is the same (One Mean)
  - Variations are by Chance (Error)
:::
- Formula: $$\hat{y} = \bar{y}$$

- Application: <br/> Everyone gets the same prediction.

## Example: Thumb Length

::: {.columns}
::: {.column width="40%"}
```{webr}
#| fig-height: 3
#| fig-width: 3

gf_boxplot(Thumb~1, data=Fingers) |>
  gf_hline(yintercept=mean(Fingers$Thumb), color='red', linetype=2)
```
:::
::: {.column width="60%" style="font-size:.8em"}
Model:

```{webr}
#| results: 'asis'
empty_model <- lm(Thumb~NULL, data=Fingers)

empty_model

```

:::
:::
<!-- end columns -->

## Empty-Model Residual

::: {.columns}
::: {.column width="40%"}
```{webr}
#| fig-height: 3
#| fig-width: 3

gf_histogram(Thumb~., data=Fingers, bins=10) |> 
  gf_boxplot(width=10) |>
  gf_hline(yintercept=mean(Fingers$Thumb), color='red', linetype=2)
```
:::
::: {.column width="60%" style="font-size:.8em"}
Residual = Observed − Predicted

```{webr}
#| fig-height: 3
#| fig-width: 3
gf_histogram(resid(empty_model)~., bins=10) |> 
  gf_boxplot(width=10) |>
  gf_hline(yintercept=0, color='blue', linetype=2)
```

:::
:::

# Modeling with Group Means
## Grouping by Gender

::: {.columns}
::: {.column}
:::{.incremental}
```{webr}
#| fig-height: 3
#| fig-width: 5
gf_dhistogram(Thumb~., data=Fingers, bins=10, fill=~Gender) |> 
  gf_boxplot(width=10, fill='yellow') |>
  #gf_hline(yintercept=gMeans, color=c('red', 'blue'), linetype=2) |>
  gf_facet_grid(.~Gender)
```
:::
:::
::: {.column}
:::{.fragment .nonincremental}
- Questions:
  - Does thumb length differ by gender?
  - Does one overall mean seem reasonable?
:::
:::
:::
<!-- end columns -->

## Model by Group (Gender)

- In word:
$$\text{Thumb} = \text{Gender} + \text{Error}$$

- In Application:
  - Each gender gets its own mean as prediction.


## Fit the Group Model

::: {.columns}
::: {.column style="font-size: .7em;" width="60%"}
```{webr}
group_model <- lm(Thumb ~ Gender, data = Fingers)
summary(group_model)
```
:::
::: {.column width="40%"}
```{webr}
#| fig-height: 3
#| fig-width: 3
gMeans=mean(Thumb~Gender, data=Fingers)
gf_boxplot(Thumb~Gender, data=Fingers, fill=~Gender) |> 
  #gf_hline(yintercept=gMeans, color=c('red', 'blue'), linetype=2) |>
  #gf_abline(intercept=gMeans[1]-diff(gMeans), slope=diff(gMeans), linewidth=1.5) |>
  gf_labs()
```
:::
:::
<!-- end columns -->

## Interpreting Coefficients
- Here, Female is the reference:
  - Intercept = mean thumb for females
  - GenderMale = difference (Male − Female)

::: {.columns}
::: {.column}
:::{.fragment}
```{webr}
mean(Thumb~Gender, data=Fingers)
```
:::
:::
::: {.column}
:::{.fragment}
```{webr}
head(predict(group_model))
```
:::
:::
:::
<!-- end columns -->


## Visualizing Predictions

```{webr}
gf_jitter(Thumb ~ Gender, data = Fingers, width = 0.1) |>
  gf_hline(yintercept=gMeans, color=c('red', 'blue'), linetype=2)
  #gf_model(group_model)
```

- Interpretation:
  * Black points: observed data
  * Red/Blue lines: predicted group means


## Compare to Empty Model

```{webr}
gf_jitter(Thumb ~ Gender, data = Fingers, width = 0.1) %>%
  gf_model(empty_model, color = "blue")
```


## Residual Comparison (Conceptual)

```{webr}
sd(resid(empty_model))
sd(resid(group_model))
```

- Discussion:

  - Which model has smaller residual spread?
  - What does that imply?


# Key Insight
:::{.nonincremental}
- The group model:
  * Allows different means
  * Reduces prediction error
  * Uses information from explanatory variable

- But…
  - Is the reduction large enough to matter?

:::
