---
title: "CourseKata Chapter 9 (9 to 11)"
subtitle: "Linear Regression"
author: "Mansour Abdoli, PhD"
format:
  live-revealjs:
    slide-number: true
    incremental: true
    chalkboard: true
    preview-links: true
    code-overflow: wrap
    footer: "CourseKata Ch. 9"
engine: knitr
execute:
  echo: true
  warning: false
  message: false
output-files: CK-ch9-sec9-11.html
---

## Overview / Goals

{{< include ./_extensions/r-wasm/live/_knitr.qmd >}}

```{r}
#| echo: false
knitr::opts_chunk$set(echo=FALSE)
suppressPackageStartupMessages({
  library(mosaic)
  library(ggformula)
  library(supernova)
})
def.par <- par(no.readonly = TRUE)
on.exit(par(def.par))

source('./assets/utils.R')
source('./assets/embedR.R')
source('./data/embedData.R')
load(file = "./data/Fingers.rda")
```

```{r}
#| echo: false
#| results: 'asis'
cat(embedR("./assets/utils.R"))
cat(embedData(Fingers))
```


```{webr}
#| echo: false
#| output: true
#| warning: false
#| message: false

library(mosaic)
library(ggformula)
library(supernova)
options(digits = 3)

.webr_ready <- TRUE
set.seed(123)
```


:::{.nonincremental}
- **Correlation (r)**
- **Correlation matrices**
- Relationship between **r and R²**
- **Null distribution of slopes**
- **Limits of regression models**

:::

## Two Numerical Variables

:::{.nonincremental}
Previously we modeled:

- response variable \(y\)
- explanatory variable \(x\)

with regression

$$
\hat{y} = b_0 + b_1x
$$

Key question:

> How strong is the relationship?

:::

## Pearson's Correlation \(r\)

Pearson’s \(r\):

- measures **strength and direction**
- of a **linear relationship**
- between **two quantitative variables**

:::{.fragment}
Range:

$$
-1 \le r \le 1
$$
:::

## Interpreting r

| r value | Meaning |
|------|------|
| close to 1 | strong positive relationship |
| close to -1 | strong negative relationship |
| close to 0 | weak or no linear relationship |

::: {.fragment}
Important:

> Correlation measures **linear association**.

:::

## Example Scatterplot

::: {.columns}
::: {.column}

```{webr}
#| fig-width: 5
#| fig-height: 4
gf_point(mpg ~ wt, data = mtcars) |>
  gf_lm()
```
:::
::: {.column}
::: {.fragment .nonincremental}
Questions:

- What direction is the relationship?
- How strong is the relationship? 
:::
:::
:::
<!-- end columns -->

## Calculating Correlation
**Pearson's r** (using R):

```{r, echo=TRUE}
cor(mtcars$wt, mtcars$mpg)
```

Interpretation:

* sign → direction
* magnitude → strength

## Correlation Matrix
**Pairwise correlations** for multiple variables:

Example:

```{r, echo=TRUE}
round(cor(mtcars[,c("mpg","wt","hp","disp")]),2)
```

- **Practice:** Interpret some the values.

## Relationship Between r and R²
::: {.columns}
::: {.column}
For **simple linear regression**:
$$R^2 = r^2$$
Where:

:::{.nonincremental}
* $r$ = correlation
* $R^2$ = proportion of variation explained
:::
:::
::: {.column}
::: {.fragment}
Example: 

$r^2:$
```{r, echo=TRUE}
cor(mtcars$wt, mtcars$mpg)^2
```

PRE:

```{r, echo=TRUE}
pre(lm(mpg ~ wt, data = mtcars))
```
:::

:::
:::
<!-- end columns -->

## Understanding the Slope

Recall the regression model:
$$\hat{y} = b_0 + b_1x$$

The slope (b_1):

> change in predicted (y) for a one-unit increase in (x).

But how do we know if this slope is meaningful?


## Null Model Idea
::: {.columns}
::: {.column}
::: {.nonincremental}
- The slope is zero: $\beta_1=0$

- **No relationship** between $x$ and $y$

- Any observed relation is due to **random variation**.
:::
:::
::: {.column}
::: {.fragment .nonincremental}
- Simulate random variation:
  1. Shuffle the response variable.
  2. Recalculate the slope.
  3. Repeat many times.
:::
:::
:::
<!-- end columns -->

## Simulation Example
::: {.columns}
::: {.column}

```{webr}
null_slopes <- do(500) * 
  b1(shuffle(mpg) ~ wt, data = mtcars)
head(null_slopes)
```
:::
::: {.column}
```{webr}
#| fig-width: 5
#| fig-height: 4
#| warning: FALSE
gf_histogram(~wt, data = null_slopes) |>
  gf_point(0~b1(mpg ~ wt, data = mtcars),
    color='red', size=2)
```
:::
:::
<!-- end columns -->

# Limitation of Regression
:::{.nonincremental}
1. **Association vs causation**
2. **Extrapolation**
3. **Linearity assumption**
:::

## Association vs Causation

Regression detects **association**, not causation.

::: {.fragment .nonincremental}

Example:

Ice cream sales ↑  
Drowning incidents ↑
:::

- Does ice cream cause drowning?

## Extrapolation
::: {.columns}
::: {.column}
Where models are reliable?

  - **within the observed data range**.

:::
::: {.column}

::: {.fragment}
Example:
```{webr}
#| fig-width: 5
#| fig-height: 4
gf_point(Petal.Length~Petal.Width, data=iris) |>
  gf_lm(Petal.Length~Petal.Width, data = iris[iris$Species=='setosa', ],
        color = "red")
```
:::
:::
:::
<!-- end columns -->

## Linearity Assumption

Linear regression assumes:

$$\text{relationship = straight line}$$

But some relationships are **curved**.

Example patterns:

* quadratic
* exponential
* saturation effects

---

## Example Nonlinear Pattern

```{r}
x <- 1:100
y <- x^2 + rnorm(length(x), 0, 100)

gf_point(y ~ x)
```

A straight line would not describe this well.


## Diagnosing Linearity
::: {.columns}
::: {.column}
::: {.nonincremental}
We check:

* scatterplots
  - Linear relation
* residual plots
  - **No systematic pattern**.

:::
:::
::: {.column}
::: {.fragment}
```{r}
x <- 50:100
y <- x^2 + rnorm(length(x), 0, 100)
mod <- lm(y~x)
gf_point(y~x) |> gf_lm()
gf_point(resid(mod)~x) |> 
  gf_abline(intercept = 0, slope=0, color='blue')
```

:::
:::
:::
<!-- end columns -->



## Key Takeaways
:::{.nonincremental}
- Correlation (r): measures strength of linear association
- Relationship: $R^2 = r^2$
- Null slope distribution:
  * helps evaluate whether slopes occur by chance
- Regression limitations:
  * association ≠ causation
  * avoid extrapolation
  * check linearity
:::