---
title: "CourseKata Chapter 12"
subtitle: "Confidence Interval (2)"
author: "Mansour Abdoli, PhD"
format:
  live-revealjs:
    slide-number: true
    incremental: true
    chalkboard: true
    preview-links: true
    code-overflow: wrap
    footer: "CourseKata Ch. 12 (2)"
engine: knitr
execute:
  echo: true
  warning: false
  message: false
output-files: CK-ch12-sec9-13.html
---

## Overview / Goals

{{< include ./_extensions/r-wasm/live/_knitr.qmd >}}

```{r}
#| echo: false
knitr::opts_chunk$set(echo=FALSE)
suppressPackageStartupMessages({
  library(mosaic)
  library(ggformula)
  library(supernova)
})
def.par <- par(no.readonly = TRUE)
on.exit(par(def.par))

theme_set(
  theme_bw() +
    theme(
    axis.text  = element_text(size = 20),
    axis.title = element_text(size = 22),
    legend.text  = element_text(size = 20),
    legend.title = element_text(size = 22)
  )
)

source('./assets/utils.R')
source('./assets/embedR.R')
source('./data/embedData.R')
load(file = "./data/TipExperiment.rda")
```

```{r}
#| echo: false
#| results: 'asis'
cat(embedR("./assets/utils.R"))
cat(embedData(TipExperiment))
```


```{webr}
#| echo: false
#| output: true
#| warning: false
#| message: false

library(mosaic)
library(ggformula)
library(supernova)
options(digits = 3)

theme_set(
  theme_bw() +
    theme(
    axis.text  = element_text(size = 20),
    axis.title = element_text(size = 22),
    legend.text  = element_text(size = 20),
    legend.title = element_text(size = 22)
  )
)

.webr_ready <- TRUE
set.seed(123)
```


:::{.nonincremental}
- **C.I. for One Mean - Review**
- **C.I. for Two Means**
- **C.I. for Intercept & Slope**
- **Paiwise C.I.**
- **Factors Affecting C.I.**
:::


# C.I. for One Mean - Review

Point Estimate $\pm$ ME

## Computing ME
:::{.columns}
:::{.column .nonincremental}
Simulation

- Resample
- Compute $\bar y$
- Repeate many times
- Find middle C%: $$(LL, UL)$$
- ME = $\frac{UL-LL}{2}$
:::
:::{.column .nonincremental}
Theoretical

- Check Assumption
  - Normality or
  - Large Sample Size
- Compute SE = $\frac{s}{\sqrt{n}}$
- Find $t_{\alpha/2}$ multiplier</br>
  from $t\sim T(df=n-1)$
- ME = $t_{\alpha/2} \cdot SE$
:::
:::

# C.I. for Two Means

Point Estimate $\pm$ ME
$$\bar y_2 - \bar y_1 \pm ME$$

## Computing ME - By Simulation
Three ways to simulate a Distribution

- Shuffle $Y$
  - Models no-association $\text{Null Distribution}$ </br>
- Resample dataset
  - Different groups sizes </br>
- Resample within groups
  - Holds the same structure  </br>

- Key is finding ME for the same structure

## ME and Sampling Distributions
:::{.columns}
:::{.column .nonincremental}
Simulation

- Resample (Find Sampling Dist.)
- Compute $b_1=\bar y_2-\bar y_1$ 
- Repeate many times
- Find middle C%: $$(LL, UL)$$
- ME = $\frac{UL-LL}{2}$
:::
:::{.column .nonincremental}
```{webr}
#| fig-width: 5
b1null<- do(1000) * 
  b1(shuffle(Tip)~Condition, data=TipExperiment)
b1boot<- do(1000) * 
  b1(Tip~Condition, data=resample(TipExperiment))
b1Gboot<- do(1000) * 
  b1(Tip~Condition, data=
  rbind(
    resample(TipExperiment |> filter(Condition=='Control')),
    resample(TipExperiment |> filter(Condition=='Smiley Face'))
  ))
sort(b1null[[1]])[c(25, 975)] # quantile(b1null[[1]], c(0.025, 0.975))
sort(b1boot$[[1]])[c(25, 975)] # quantile(b1boot$[[1]], c(0.025, 0.975))
sort(b1Gboot[[1]])[c(25, 975)] # quantile(b1Gboot$[[1]], c(0.025, 0.975))

```

:::
:::

## Computing ME - Theoretical
:::{style="font-size: .9em;"}

- Check Assumptions
:::
:::{.columns  .nonincremental}
:::{.column style="font-size: .7em;"}

  - Normality or
:::
:::{.column style="font-size: .7em;"}

  - Large Sample Sizes
:::
:::

:::{style="font-size: .9em;"}

- Compute SE & DF
:::

:::{.columns style="font-size: .7em;"}
:::{.column}

- Equal Variance
  - $s_p^2 = \frac{(n_1)s_1^2+(n_2)s_2^2}{n_1+n_2-2}$
  - $DF = n_1+n_2-2$
  - SE = $s_p\sqrt{\frac{1}{n_1}+\frac{1}{n_2}}$
:::
:::{.column}

- Unequal Variance
  - $DF \approx min(n_1-1, n_2-1)$
  - SE = $\sqrt{\frac{s_1^2}{n_1}+\frac{s_2^2}{n_2}}$
:::
:::

:::{ style="font-size: .9em;"}

- $t_{\alpha/2}$ from $t\sim T(df=DF)$
- ME = $t_{\alpha/2} \cdot SE$
:::

## Computing ME - Theoretical
'confint()' Computes confidence interval

```{webr}
TipExperiment |> group_by(Condition) |> 
  summarise(n=n(), Mean=mean(Tip), SD=sd(Tip))

lmdl<-lm(Tip~Condition, data=TipExperiment)
# Summary table provides SE
#summary(lmdl)

#confint(lmdl)

```

# C.I. for Intercept & Slope

## C.I. for Intercept & Slope
Simulation

:::{.columns}
:::{.column width="24%"}
- Resample Data
:::
:::{.column width="38%"}
- Compute $b_0$ and $b_1$
:::
:::{.column width="38%"}
- Repeate many times
:::
:::

:::{.columns}
:::{.column}
- For $\beta_0$:
  - Find middle C%: $$(LL_{b_0}, UL_{b_0})$$
  - $ME_{b_0} = \frac{UL_{b_0}-LL_{b_0}}{2}$
  - $b_0\pm ME_{b_0}$
:::
:::{.column}
:::{.fragment .nonincremental}
- For $\beta_1$:
  - Find middle C%: $$(LL_{b_1}, UL_{b_1})$$
  - $ME_{b_1} = \frac{UL_{b_1}-LL_{b_1}}{2}$
  - $b_1\pm ME_{b_1}$
:::
:::
:::

## C.I. for Intercept & Slope
Theoretical

:::{.columns}
:::{.column .nonincremental}

- Assumptions:
  - Linear Association
  - Normal Distribution
  - Equal Variance

- $t_{\alpha/2}$ from $t\sim T(df=n-2)$
- $ME_{b_i} = t_{\alpha/2} \cdot SE_{b_i}$

:::
:::{.column}
- $b_i$ and $SE_i$ are from the Summary table

:::{.fragment style="font-size: .5em;"}


:Coefficients 

|            | Estimate | Std. Error | t value | Pr(>\|t\|) |
|------------|----------|------------|---------|----------|
| (Intercept)|  $b_0$   | $SE_{b_0}$  |$t_{b_0}=\frac{b_0}{SE_{b_0}}$| $p\text{-value}_{b_0}$|
| Explanatory|  $b_1$   | $SE_{b_1}$  |$t_{b_0}=\frac{b_1}{SE_{b_1}}$| $p\text{-value}_{b_1}$|

:::
:::
:::

## C.I. for Intercept & Slope
Recalculate C.I.s:

```{webr}
lmdl<- lm(Tip~Check, data=TipExperiment)
summary(lmdl)
confint(lmdl)
```


# Paiwise C.I.
Adhoc for Single Test of Multiple Predictors

## Account for $\alpha$ Inflation

- Benferroni:
  - Use $\alpha/K$
    - $K$ is the number of pairwise tests

- Tukey's HSD
  - Use range distribution

:::{.fragment}
```{webr}
# A quick 3-group example for pairwise CIs
pair_dat <- data.frame(
  wait_time = c(15, 14, 16, 13, 15, 14,
                12, 11, 13, 12, 10, 11,
                 9,  8, 10,  9,  8,  9),
  method = factor(rep(c("Standard", "Express", "Kiosk"), each = 6))
)

pair_fit <- aov(wait_time ~ method, data = pair_dat)
summary(pair_fit)
TukeyHSD(pair_fit)
```
:::


# Factors Affecting C.I.

## Main Ideas

:::{.nonincremental}
- Higher confidence level $\Rightarrow$ wider interval
- Larger sample size $\Rightarrow$ narrower interval
- More variability $\Rightarrow$ wider interval
- For two means: more balanced group sizes $\Rightarrow$ smaller SE
- For regression slope: more spread in $X$ $\Rightarrow$ narrower interval for $b_1$
:::

## A Quick Demo

```{webr}
ci_width_mean <- function(n = 25, sd = 10, level = .95) {
  se <- sd / sqrt(n)
  t_star <- qt((1 + level)/2, df = n - 1)
  2 * t_star * se
}

data.frame(
  setting = c("baseline", "bigger n", "higher sd", "higher level"),
  width = c(
    ci_width_mean(n = 25, sd = 10, level = .95),
    ci_width_mean(n = 100, sd = 10, level = .95),
    ci_width_mean(n = 25, sd = 20, level = .95),
    ci_width_mean(n = 25, sd = 10, level = .99)
  )
)
```

## Wrap-Up Questions

:::{.nonincremental}
- If we want a narrower interval, which things can we control?
- Why does asking for 99% confidence force a wider interval?
- Why does larger sample size help even when the center stays the same?
- Why is a slope estimate more precise when the $X$ values are more spread out?
:::