---
title: "CourseKata Chapter 11"
subtitle: "Model Significance (1)"
author: "Mansour Abdoli, PhD"
format:
  live-revealjs:
    slide-number: true
    incremental: true
    chalkboard: true
    preview-links: true
    code-overflow: wrap
    footer: "CourseKata Ch. 11"
engine: knitr
execute:
  echo: true
  warning: false
  message: false
output-files: CK-ch11-sec1-9.html
---

## Overview / Goals

{{< include ./_extensions/r-wasm/live/_knitr.qmd >}}

```{r}
#| echo: false
knitr::opts_chunk$set(echo=FALSE)
suppressPackageStartupMessages({
  library(mosaic)
  library(ggformula)
  library(supernova)
})
def.par <- par(no.readonly = TRUE)
on.exit(par(def.par))

theme_set(
  theme_bw() +
    theme(
    axis.text  = element_text(size = 20),
    axis.title = element_text(size = 22),
    legend.text  = element_text(size = 20),
    legend.title = element_text(size = 22)
  )
)

source('./assets/utils.R')
source('./assets/embedR.R')
source('./data/embedData.R')
load(file = "./data/Fingers.rda")
```

```{r}
#| echo: false
#| results: 'asis'
cat(embedR("./assets/utils.R"))
cat(embedData(Fingers))
```


```{webr}
#| echo: false
#| output: true
#| warning: false
#| message: false

library(mosaic)
library(ggformula)
library(supernova)
options(digits = 3)

theme_set(
  theme_bw() +
    theme(
    axis.text  = element_text(size = 20),
    axis.title = element_text(size = 22),
    legend.text  = element_text(size = 20),
    legend.title = element_text(size = 22)
  )
)

.webr_ready <- TRUE
set.seed(123)
```


:::{.nonincremental}
- **Assumption $\rightarrow$ Null Dist. $\rightarrow$ Observation**
- **Theoretical Null Distribution**
- **Test of Significance: $\beta_1$**
- **Type I and Type II Errors**
- **Complex v. Simple Models**
- **Type I Error Inflation**

- **PRE Theoretical Distributions**
- **F Theoretical Distributions**
- **Pairwise Comparisons**
- **Test of Independence**

:::


# Assumption $\rightarrow$ Null Dist. $\rightarrow$ Observation

## Sampling Variability

* Different samples → different statistics
* Statistics **vary naturally**

:::{.fragment}
Key idea:

> Variation is normal, not suspicious
:::

## Sampling Distribution

* Distribution of a statistic across many samples
* Built by:
  * Re-sampling OR
  * Mathematical theory

## Null Distribution
**Null Hypothesis ($H_0$):**

:::{.columns}
:::{.column}
:::{.fragment .nonincremental}
* Assumption:
  * No effect
  * No difference
  * No relationship
:::
:::
:::{.column}
* Null Distribution of $b_1$:
  * Categorical Explanatory
    * Mean difference = 0 
  * Quantitative Explanatory
    * Slope = 0
:::
:::

## Null Distribution: Simulation
:::{.columns}
:::{.column}

* Simulate no Association:
  * Shuffle Outcome
* Compute $b_1$
* Repeat many times
* Visualize (Histogram)
* Add the observed value
:::
:::{.column}
:::{.fragment}
```{webr}
#| fig-width: 5
#| fig-heigh: 3
sdob1 <- do(500) *
  b1(shuffle(Thumb)~Gender, data=Fingers) |> 
  setNames('b1')
gf_dhistogram(~b1, data = sdob1, binwidth=1) |>
  gf_point(0~x, data=data.frame(x=b1(Thumb~Gender, data=Fingers)), color='red', size=3)
```
:::
:::
:::

# Theotetical Null Distribution
## Null Distribution: Mathematical Model

:::{.fragment .nonincremental}

- Assumption: Error Distribution $\sim N(0, \sigma)$

::: 
 
:::{.columns}
:::{.column .fragment .nonincremental}
Categorical Explanatory

  * Normal
  * Mean of 0
  * Same group variances  
:::
:::{.column .fragment .nonincremental}
Quantitative Explanatory

  * Normal
  * Mean of 0
  * Variance constant across $X$
:::
:::

- Null Distribution: $\frac{b_1}{s_{b1}}\sim t(df=n-2)$

## Example 1: Categorical Explanatory
$$Y_A\sim N(20, 3), \quad Y_B\sim N(20, 3), \quad n_1=n_2=10$$

:::{.columns}
:::{.column width="33%"}
```{r}
#| fig-width: 3
#| fig-height: 5
set.seed(123)
mu<- c(20, 20); sd<- c(1, 1)*3; n<-c(1, 1)*10
skw <- c(1, 1)
w<- 0.5 # binwidth
dset<- data.frame(
    x=c(rep('A', n[1]), rep('B', n[2])),
    y=c(rskw(n[1], mu[1], sd[1], skw),
        rskw(n[2], mu[2], sd[2], 1)
        ))
gf_histogram(~y, data=dset, binwidth=w) |>
  gf_facet_grid(x~.)
```
:::
:::{.column width="33%"}
```{r}
#| fig-width: 3
#| fig-height: 5
gf_qq(~y, data=dset) |> gf_qqline(color='blue') |>
  gf_facet_grid(~x)
```
:::
:::{.column width="33%"}
```{r}
#| fig-width: 3
#| fig-height: 5
gf_boxplot(y~x, data=dset)    
```
:::
:::

## Example 1: Null Distributions

```{r}
  sdob1 <- do(1000) *
    b1(shuffle(y)~x, data=dset) |>
    setNames('b1')
  b1.bar <- mean(sdob1$b1)
  b1.sd <- sd(sdob1$b1)
  n<- length(sdob1$b1)
  x.lim<- b1.bar+c(-1,1)*4*b1.sd

  b1.qts<- quantile(sdob1$b1, c(.025, 0.975))
  th.qts<- b1.bar+b1.sd*qt(c(.025, 0.975), df=n-2)
  sdob1$tail<- factor(sdob1$b1<b1.qts[1] | sdob1$b1>b1.qts[2])
  gf_histogram(~b1, data = sdob1, binwidth=w, fill=~tail, show.legend = FALSE) |>
    gf_function(fun=b1dt, args=list(df=n-2, b1=sdob1$b1, w=w),xlim=x.lim)|>
    gf_fun_fill(fun=function(x) b1dt(x, df=n-2, b1=sdob1$b1, w=w), from = x.lim[1], to = th.qts[1], color='red') |>
    gf_fun_fill(fun=function(x) b1dt(x, df=n-2, b1=sdob1$b1, w=w), from=th.qts[2], to=x.lim[2], color='red')
```

## Example (2): Different Variance
$$Y_A\sim N(20, 6), \quad Y_B\sim N(20, 3), \quad n_1=n_2=10$$

:::{.columns}
:::{.column}
```{r}
#| fig-width: 5
#| fig-height: 2.5
set.seed(123)
mu<- c(20, 20); sd<- c(2, 1)*3; n<-c(1, 1)*10
skw <- c(1, 1)
w<- 0.5 # binwidth
dset<- data.frame(
    x=c(rep('A', n[1]), rep('B', n[2])),
    y=c(rskw(n[1], mu[1], sd[1], skw),
        rskw(n[2], mu[2], sd[2], 1)
        ))
gf_qq(~y, data=dset) |> gf_qqline(color='blue') |>
  gf_facet_grid(~x)
```
```{r}
#| fig-width: 5
#| fig-height: 2.5
gf_boxplot(x~y, data=dset)    
```
:::
:::{.column}
```{r}
#| fig-width: 5
#| fig-height: 5
sdob1 <- do(1000) *
  b1(shuffle(y)~x, data=dset) |>
  setNames('b1')
b1.bar <- mean(sdob1$b1)
b1.sd <- sd(sdob1$b1)
n<- length(sdob1$b1)
x.lim<- b1.bar+c(-1,1)*4*b1.sd

b1.qts<- quantile(sdob1$b1, c(.025, 0.975))
th.qts<- b1.bar+b1.sd*qt(c(.025, 0.975), df=n-2)
sdob1$tail<- factor(sdob1$b1<b1.qts[1] | sdob1$b1>b1.qts[2])
gf_histogram(~b1, data = sdob1, binwidth=w, fill=~tail, show.legend = FALSE) |>
  gf_function(fun=b1dt, args=list(df=n-2, b1=sdob1$b1, w=w),xlim=x.lim)|>
  gf_fun_fill(fun=function(x) b1dt(x, df=n-2, b1=sdob1$b1, w=w), from = x.lim[1], to = th.qts[1], color='red') |>
  gf_fun_fill(fun=function(x) b1dt(x, df=n-2, b1=sdob1$b1, w=w), from=th.qts[2], to=x.lim[2], color='red')
```
:::
:::

## Example (3): Unbalanced Data
$$Y_A\sim N(20, 6), \quad Y_B\sim N(20, 3), \quad n_1=10, n_2=15$$

:::{.columns}
:::{.column}
```{r}
#| fig-width: 5
#| fig-height: 2.5
set.seed(123)
mu<- c(20, 20); sd<- c(2, 1)*3; n<-c(1, 1.5)*10
skw <- c(1, 1)
w<- 0.5 # binwidth
dset<- data.frame(
    x=c(rep('A', n[1]), rep('B', n[2])),
    y=c(rskw(n[1], mu[1], sd[1], skw),
        rskw(n[2], mu[2], sd[2], 1)
        ))
gf_qq(~y, data=dset) |> gf_qqline(color='blue') |>
  gf_facet_grid(~x)
```
```{r}
#| fig-width: 5
#| fig-height: 2.5
gf_boxplot(x~y, data=dset)    
```
:::
:::{.column}
```{r}
#| fig-width: 5
#| fig-height: 5
sdob1 <- do(1000) *
  b1(shuffle(y)~x, data=dset) |>
  setNames('b1')
b1.bar <- mean(sdob1$b1)
b1.sd <- sd(sdob1$b1)
n<- length(sdob1$b1)
x.lim<- b1.bar+c(-1,1)*4*b1.sd

b1.qts<- quantile(sdob1$b1, c(.025, 0.975))
th.qts<- b1.bar+b1.sd*qt(c(.025, 0.975), df=n-2)
sdob1$tail<- factor(sdob1$b1<b1.qts[1] | sdob1$b1>b1.qts[2])
gf_histogram(~b1, data = sdob1, binwidth=w, fill=~tail, show.legend = FALSE) |>
  gf_function(fun=b1dt, args=list(df=n-2, b1=sdob1$b1, w=w),xlim=x.lim)|>
  gf_fun_fill(fun=function(x) b1dt(x, df=n-2, b1=sdob1$b1, w=w), from = x.lim[1], to = th.qts[1], color='red') |>
  gf_fun_fill(fun=function(x) b1dt(x, df=n-2, b1=sdob1$b1, w=w), from=th.qts[2], to=x.lim[2], color='red')
```
:::
:::

## Example (4): Skewed Distribution
$$Y_A\sim N(20, 3), \quad Y_B\sim \text{skewed right}, \quad n_1=10, n_2=15$$

:::{.columns}
:::{.column}
```{r}
#| fig-width: 5
#| fig-height: 2.5
set.seed(123)
mu<- c(20, 20); sd<- c(1, 1)*3; n<-c(1, 1.5)*10
skw <- c(1, 5)
w<- 0.5 # binwidth
dset<- data.frame(
    x=c(rep('A', n[1]), rep('B', n[2])),
    y=c(rskw(n[1], mu[1], sd[1], skw),
        rskw(n[2], mu[2], sd[2], 1)
        ))
gf_qq(~y, data=dset) |> gf_qqline(color='blue') |>
  gf_facet_grid(~x)
```
```{r}
#| fig-width: 5
#| fig-height: 2.5
gf_boxplot(x~y, data=dset)    
```
:::
:::{.column}
```{r}
#| fig-width: 5
#| fig-height: 5
sdob1 <- do(1000) *
  b1(shuffle(y)~x, data=dset) |>
  setNames('b1')
b1.bar <- mean(sdob1$b1)
b1.sd <- sd(sdob1$b1)
n<- length(sdob1$b1)
x.lim<- b1.bar+c(-1,1)*4*b1.sd

b1.qts<- quantile(sdob1$b1, c(.025, 0.975))
th.qts<- b1.bar+b1.sd*qt(c(.025, 0.975), df=n-2)
sdob1$tail<- factor(sdob1$b1<b1.qts[1] | sdob1$b1>b1.qts[2])
gf_histogram(~b1, data = sdob1, binwidth=w, fill=~tail, show.legend = FALSE) |>
  gf_function(fun=b1dt, args=list(df=n-2, b1=sdob1$b1, w=w),xlim=x.lim)|>
  gf_fun_fill(fun=function(x) b1dt(x, df=n-2, b1=sdob1$b1, w=w), from = x.lim[1], to = th.qts[1], color='red') |>
  gf_fun_fill(fun=function(x) b1dt(x, df=n-2, b1=sdob1$b1, w=w), from=th.qts[2], to=x.lim[2], color='red')
```
:::
:::

## Example (5): Skewed & large $n$
$$Y_A\sim N(20, 3), \quad Y_B\sim \text{skewed right}, \quad n_1=100, n_2=150$$


:::{.columns}
:::{.column}
```{r}
#| fig-width: 5
#| fig-height: 2.5
set.seed(123)
mu<- c(20, 20); sd<- c(1, 1)*3; n<-c(1, 1.5)*100
skw <- c(1, 5)
w<- 0.5 # binwidth
dset<- data.frame(
    x=c(rep('A', n[1]), rep('B', n[2])),
    y=c(rskw(n[1], mu[1], sd[1], skw),
        rskw(n[2], mu[2], sd[2], 1)
        ))
gf_qq(~y, data=dset) |> gf_qqline(color='blue') |>
  gf_facet_grid(~x)
```
```{r}
#| fig-width: 5
#| fig-height: 2.5
gf_boxplot(x~y, data=dset)    
```
:::
:::{.column}
```{r}
#| fig-width: 5
#| fig-height: 5
sdob1 <- do(1000) *
  b1(shuffle(y)~x, data=dset) |>
  setNames('b1')
b1.bar <- mean(sdob1$b1)
b1.sd <- sd(sdob1$b1)
n<- length(sdob1$b1)
x.lim<- b1.bar+c(-1,1)*4*b1.sd

b1.qts<- quantile(sdob1$b1, c(.025, 0.975))
th.qts<- b1.bar+b1.sd*qt(c(.025, 0.975), df=n-2)
sdob1$tail<- factor(sdob1$b1<b1.qts[1] | sdob1$b1>b1.qts[2])
gf_histogram(~b1, data = sdob1, binwidth=w, fill=~tail, show.legend = FALSE) |>
  gf_function(fun=b1dt, args=list(df=n-2, b1=sdob1$b1, w=w),xlim=x.lim)|>
  gf_fun_fill(fun=function(x) b1dt(x, df=n-2, b1=sdob1$b1, w=w), from = x.lim[1], to = th.qts[1], color='red') |>
  gf_fun_fill(fun=function(x) b1dt(x, df=n-2, b1=sdob1$b1, w=w), from=th.qts[2], to=x.lim[2], color='red')
```
:::
:::

## Null Distribution Summary
For **binary** of **numeriacl** $X$:
$$Y=\beta_0+\beta_1 X + \varepsilon$$

- If assumptions met: Error Distribution $\sim N(0, \sigma)$
  - $\frac{b_1-\beta_1}{s_{b_1}} \sim t(df=n-2)$

:::{style="font-size:.7em;"}
  - $t(df)$:
    - a density centered at $0$
    - more spread than $Z$
    - as $df (n)\to \infty, \quad t\xrightarrow{D} z$
:::

# Test of Significance $\beta_1$

## $H_a$ & Unusual (Unexpected) Event
- Assumption & $H_0 \to $ Null Distribution
- Alternative:
  - $\beta_1>0$: Reject $H_0$ if too far to the right.
  - $\beta_1<0$: Reject $H_0$ if too far to the left.
  - $\beta_1\ne0$: Reject $H_0$ if too far from the middle (expected).

## Thumb~Gender Example
$$Thumb=\beta_0+\beta_1 Gender_{Male} + \varepsilon$$
- $H_0: \beta_1=0$ vs. $H_a: \beta_1\ne 0$

:::{.columns}
:::{.column}
:::{.fragment}
```{r}
#| fig-width: 5
#| fig-heigh: 3
sdob1 <- do(500) *
  b1(shuffle(Thumb)~Gender, data=Fingers)  |>
  setNames('b1')
b1.bar <- mean(sdob1$b1)
b1.sd <- sd(sdob1$b1)
n<- length(sdob1$b1)
obs_b1 <- b1(Thumb~Gender, data=Fingers)
x.lim<- b1.bar+c(-1,1)*4*b1.sd
w<- 0.5 # binwidth

b1.qts<- quantile(sdob1$b1, c(.025, 0.975))
th.qts<- b1.bar+b1.sd*qt(c(.025, 0.975), df=n-2)
sdob1$tail<- factor(sdob1$b1<b1.qts[1] | sdob1$b1>b1.qts[2])
gf_histogram(~b1, data = sdob1, binwidth=w, fill=~tail, show.legend = FALSE) |>
  gf_function(fun=b1dt, args=list(df=n-2, b1=sdob1$b1, w=w),xlim=x.lim)|>
  gf_fun_fill(fun=function(x) b1dt(x, df=n-2, b1=sdob1$b1, w=w), from = x.lim[1], to = th.qts[1], color='red') |>
  gf_fun_fill(fun=function(x) b1dt(x, df=n-2, b1=sdob1$b1, w=w), from=th.qts[2], to=x.lim[2], color='red') |>
  gf_point(0~b1, data=data.frame(b1=obs_b1, tail=TRUE), color='red', size=3, show.legend=FALSE) |>
  gf_labs(title=expression('Rejection Region(' * alpha==0.05 * ')'))
```
:::
:::
:::{.column}
:::{.fragment}
```{r}
#| fig-width: 5
#| fig-heigh: 3
sdob1$pval.tail<- factor(abs(sdob1$b1)>abs(obs_b1))
gf_histogram(~b1, data = sdob1, binwidth=w, fill=~pval.tail, show.legend = FALSE) |>
  gf_point(0~b1, data=data.frame(b1=obs_b1, pval.tail=TRUE), color='red', size=3, show.legend=FALSE) |>
  gf_labs(title='p-value')
```

:::
:::
:::

> Probability of getting a result this extreme (or more)
> **if the null is true**

## Decision
Reject $H_0$, at $\alpha$ significance level, if:

:::{.columns}
:::{.column}
- Observed $b_1$ is in Rejection Region
:::
:::{.column}
- $p$-value is less than $\alpha$
:::
:::

:::{.fragment}
otherwise,
:::
- Fail to reject $H_0$
- Both $H_a$ and $H_0$ are plausible.

# Type I & II Errors


## Type I Error

Rejecting a **true** null

:::{.columns}
:::{.column .fragment .nonincremental}
- False Positive
- False Alarm
:::
:::{.column .fragment .nonincremental}
$$\begin{align*}\alpha &= P(\text{Rejecting } H_0|H_0) \\
\\
&=\frac{FP}{FP+TN}\end{align*}$$
:::
:::

---

## Type II Error

Failing to reject a **false** null

:::{.columns}
:::{.column .fragment .nonincremental}
- False Negative
- Missed Opportunity
:::
:::{.column .fragment .nonincremental}
$$\begin{align*}\beta &= P(\text{Not Rejecting } H_0|H_a) \\
\\
&=\frac{FN}{FN+TP}\end{align*}$$
:::
:::

## Full Workflow

1. State null
2. Compute statistic
3. Find null distribution
4. Compute p-value
5. Decide

---

## 🧪 Slide 17 — One Full Example (IMPORTANT)

Use:

* Mean test OR slope test

Walk through:

* Null: slope = 0
* Observed slope
* Null distribution
* p-value
* Decision

👉 This is the anchor of the session

---

## 🔍 Slide 18 — Assumptions (Light Touch)

* Independence
* Approximate normality (or large n)

👉 Why:

> Needed for mathematical null distributions

---

## 🔄 Slide 19 — Simulation vs Mathematical

* Simulation = flexible, intuitive
* Mathematical = faster, standard

👉 Same logic underneath

---

## 🎯 Slide 20 — Takeaway

> All inference = comparing data to a world where nothing is happening

---

---

# 📗 **Session 2 Slides: Extending the Framework**

## 🎯 Slide 1 — Title

**One Logic, Many Tests**

---

## 🔁 Slide 2 — Review

* Null hypothesis
* Null distribution
* P-value

👉 Same process every time

---

## 🧠 Slide 3 — Big Idea

> Different tests = different statistics
> Same logic = always compare to null

---

## 📊 Slide 4 — Multiple Groups (ANOVA Idea)

* Compare more than 2 means

Null:

> All group means are equal

---

## ⚖️ Slide 5 — ANOVA Statistic

* Measures:

  * Between-group variability
  * Relative to within-group variability

👉 Big value → evidence against null

---

## 📈 Slide 6 — Regression Test

Null:

> Slope ($b_1$) = 0

👉 No relationship

---

## 📊 Slide 7 — Regression Logic

* Compare observed slope
* To slopes expected by chance

👉 Same workflow as Session 1

---

## 🧮 Slide 8 — Chi-Square (New Context)

* Two categorical variables

Null:

> Variables are independent

---

## 📉 Slide 9 — Chi-Square Idea

* Compare:

  * Observed counts
  * Expected counts under independence

---

## 📊 Slide 10 — Shape of χ² Distribution

* Not symmetric
* Right-skewed

👉 Important conceptual shift

---

## 🎯 Slide 11 — Unifying Slide (VERY IMPORTANT)

| Test       | Statistic  | Null          |
| ---------- | ---------- | ------------- |
| Mean       | difference | no difference |
| Regression | slope      | slope = 0     |
| ANOVA      | F          | equal means   |
| Chi-square | χ²         | independence  |

👉 Same logic across all

---

## 🚨 Slide 12 — Type I Error Reminder

* False positive

---

## ⚠️ Slide 13 — Type I Error Inflation

* More tests → more false positives

---

## 📊 Slide 14 — Example

If α = 0.05:

* 1 test → 5% chance false positive
* 20 tests → much higher chance

---

## 🧠 Slide 15 — Why This Matters

* Real-world research
* Multiple comparisons

---

## 🛠️ Slide 16 — Practical Thinking

* Be cautious with many tests
* Look for patterns, not isolated results

---

## 🔁 Slide 17 — Full Circle

> Same 5-step process:

1. Null
2. Statistic
3. Null distribution
4. P-value
5. Decision

---

## 🎯 Slide 18 — Final Takeaway

> Inference is not about formulas
> It’s about comparing reality to randomness

---

---

# 💡 Optional Add-Ons (Highly Recommended)

### In-class activities

* **Session 1:** simulate null distribution (shuffle slope)
* **Session 2:** match test → null → statistic activity

-->