---
title: "CourseKata Chapter 13"
subtitle: "Multivariate Models Inference"
author: "Mansour Abdoli, PhD"
format:
  live-revealjs:
    slide-number: true
    incremental: true
    chalkboard: true
    preview-links: true
    code-overflow: wrap
    footer: "CourseKata Ch. 13 (2)"
engine: knitr
execute:
  echo: true
  warning: false
  message: false
output-files: CK-ch13-sec7-9.html
---

## Overview / Goals

{{< include ./_extensions/r-wasm/live/_knitr.qmd >}}

```{r}
#| echo: false
knitr::opts_chunk$set(echo=FALSE)
suppressPackageStartupMessages({
  library(mosaic)
  library(ggformula)
  library(supernova)
})
def.par <- par(no.readonly = TRUE)
on.exit(par(def.par))

theme_set(
  theme_bw() +
    theme(
    axis.text  = element_text(size = 20),
    axis.title = element_text(size = 22),
    legend.text  = element_text(size = 20),
    legend.title = element_text(size = 22)
  )
)

source('./assets/utils.R')
source('./assets/embedR.R')
source('./data/embedData.R')
load(file = "./data/Smallville.rda")
```

```{r}
#| echo: false
#| results: 'asis'
cat(embedR("./assets/utils.R"))
cat(embedData(Smallville))
```


```{webr}
#| echo: false
#| output: true
#| warning: false
#| message: false

library(mosaic)
library(ggformula)
library(supernova)
options(digits = 3)

theme_set(
  theme_bw() +
    theme(
    axis.text  = element_text(size = 20),
    axis.title = element_text(size = 22),
    legend.text  = element_text(size = 20),
    legend.title = element_text(size = 22)
  )
)

.webr_ready <- TRUE
set.seed(123)
```


:::{.nonincremental}
- Review: Multivariate Model
  - Interpreting Coefficients
  - Comparing Models
- New Topic: Is the model real (significant)?
  - Simulated Null Distribution:
    - Is PRE/F/$\beta_j$ real?
  - Theoretical Null Distribution
    - Assumptions
    - F/$b_j$ Distribution
:::

## Multivariate Model

$$
\text{PriceK} = b_0 + b_1(\text{HomeSizeK}) + b_2(\text{Neighborhood})
$$

```{webr}
multi_model <- lm(PriceK~HomeSizeK + Neighborhood, data=Smallville)
coef(multi_model)
```


## Interpreting Coefficeints

```{r}
multi_model <- lm(PriceK~HomeSizeK + Neighborhood, data=Smallville)
coef(multi_model)
```

- $b_0=`r round(coef(multi_model)[1], 2)`$: Price in \$1000 at baseline (Downtown & 0 sq.ft)
- $b_1=`r round(coef(multi_model)[2], 2)`$: Difference in Price in \$1000 for any increase of 1000 sq.ft. in home size in the same neighborhood.
- $b_2=`r round(coef(multi_model)[3], 2)`$: Difference in Price in \$1000 of the same home size between Downtown and Eastside.

## Comparing Models

::: {.columns}
::: {.column width="25%"}
```{r}
#| fig-width: 2.5
#| fig-height: 3.5
null_model <- lm(PriceK~NULL, data=Smallville)
gf_jitter(PriceK~'Overall', data=Smallville, width=.2) |> gf_labs(x='') |> 
  gf_model(null_model)
cat('PRE=', pre(null_model), '\n')
```
:::
::: {.column width="25%"}

```{r}
#| fig-width: 2.5
#| fig-height: 3.5
Neighborhood_model <- lm(PriceK ~ Neighborhood, data = Smallville)
gf_jitter(PriceK~Neighborhood, data=Smallville, width=.15, color=~Neighborhood, show.legend=FALSE) |> 
  gf_model(Neighborhood_model)
cat('PRE=', pre(Neighborhood_model))
```

:::
::: {.column width="25%"}

```{r}
#| fig-width: 2.5
#| fig-height: 3.5
HomeSizeK_model <- lm(PriceK ~ HomeSizeK, data = Smallville)
gf_point(PriceK~HomeSizeK, data=Smallville) |> 
  gf_model(HomeSizeK_model)
cat('PRE=', pre(HomeSizeK_model))
```

:::
::: {.column width="25%"}

```{r}
#| fig-width: 2.5
#| fig-height: 3.5
Smallville$multi_predict <- predict(multi_model)
gf_point(PriceK ~ HomeSizeK, data = Smallville,
         color = ~ Neighborhood, show.legend=FALSE) %>%
  gf_lm(multi_predict ~ HomeSizeK, color = ~ Neighborhood, show.legend=FALSE)  #gf_model(multi_model)
cat('PRE=', pre(multi_model))
```

:::
:::

# Is the model real (significant)?

## Hypotheses & Null Distribution
- Null Hypothesis:
  - No; there is no different
- Alternative Hypothesis:
  - Yes; the difference is not by chance

\ 

- Build Null Distribution
- Compare with Observed statistic

# Inference by Simulation

## Is PRE difference real?
```{webr}
#| fig-height: 3.5
obs_pre <- pre(multi_model)

ndo_pre <- replicate(1000, pre(
  lm(shuffle(PriceK) ~ HomeSizeK + Neighborhood, data = Smallville)
))

gf_histogram(~ndo_pre, bins = 25) |>
  gf_vline(xintercept = obs_pre, col='red')
```

- is it?

## Is F difference real?
```{webr}
#| fig-height: 3.5
obs_f <- f(multi_model)

ndo_f <- replicate(1000, f(
  lm(shuffle(PriceK) ~ HomeSizeK + Neighborhood, data = Smallville)
))

gf_histogram(~ndo_f, bins = 25) |>
  gf_vline(xintercept = obs_f, col='red')
```

- is it?

## Are slopes meaningfully different from 0?
```{webr}
#| fig-height: 3.5
obs_betas <- coef(multi_model)[]

ndo_betas <- replicate(1000, coef(
  lm(shuffle(PriceK) ~ HomeSizeK + Neighborhood, data = Smallville)
)) |> t() |> as.data.frame()

ndo_betas |> 
  tidyr::pivot_longer(everything(), 
    names_to='b', values_to='estimate') |>
  gf_histogram(~estimate, bins=30) |>
    gf_facet_grid(b~.) |>
    gf_vline(xintercept=~estimate,
      data=data.frame(b=colnames(ndo_betas), 
      estimate=obs_betas), color='red') + theme(text=element_text(size=28))
```

## Theoretical Distribution of $F$ & $b_j$

- Population Parameters: $\beta_0$, $\beta_1$, $\beta_2$  
- Sample Statistics: $b_0$, $b_1$, $b_2$

- Null Distribution Assumptions: 
  - Observations: Independent (Within & Between)
  - Error Distribution:
    - (Roughly) Normal 
    - Constant $\sigma$

## Checking Normality/Constant Variance:

::: {.columns}
::: {.column}
```{webr}
es <- resid(multi_model)
gf_histogram(~es, bins=30)

```
:::
::: {.column}
```{webr}
gf_qq(~es) |> gf_qqline()
```

:::
:::

## F Inference

- NULL Hypothesis: All slopes are zero
  - No association beteen response and predictors.

- Null Distribution: $F\sim F(df_1=2, df_2=n-3)$
  - $p$-value $= P(F>F_{obs})$  

```{webr}
supernova(multi_model)
```

## $\beta_j$ Inference
$$
t = \frac{b_j}{SE(b_j)}
$$

- NULL Hypothesis: usually $\beta_j=0$
  - $j^{th}$ predictor has no effect.

- Null Distribution: $t\sim T(df=n-3)$
  - $p$-value $= P(t \text{ at least as extrem as } t_{obs} \text{ in favor of } H_a)$  

```{webr}
summary(multi_model)
```

## Practice
```{r}
summary(multi_model) -> mdlSummary
tTable <- mdlSummary$coefficients
tTable
```

Does `NeighborhoodEastside` have a negative impact on `PriceK`?

- Hypotheses: $H_0: \beta_{Neighborhood} = 0$ vs. $H_a: \beta_{Neighborhood} < 0$
- $p$-value = P(T<`r tTable["NeighborhoodEastside", "t value"]`) = ?

## Practice
```{r}
tTable
```

Find a 95% C.I. for $\beta_\text{HomeSizeL}$: 

- SE $= `r tTable["HomeSizeK", "Std. Error"]`
- $t^* = t_{0.025, df=`r multi_model$df.residual`} = `r -round(qt(.025, multi_model$df.residual),4)`$
- 95% C.I. of $\beta_{\text{HomeSizeK}} = (...., ....)$


