---
title: "CourseKata Chapter 11"
subtitle: "Model Significance (2)"
author: "Mansour Abdoli, PhD"
format:
  live-revealjs:
    slide-number: true
    incremental: true
    chalkboard: true
    preview-links: true
    code-overflow: wrap
    footer: "CourseKata Ch. 11 (2)"
engine: knitr
execute:
  echo: true
  warning: false
  message: false
output-files: CK-ch11-sec11-13.html
---

## Overview / Goals

{{< include ./_extensions/r-wasm/live/_knitr.qmd >}}

```{r}
#| echo: false
knitr::opts_chunk$set(echo=FALSE)
suppressPackageStartupMessages({
  library(mosaic)
  library(ggformula)
  library(supernova)
})
def.par <- par(no.readonly = TRUE)
on.exit(par(def.par))

source('./assets/utils.R')
source('./assets/embedR.R')
source('./data/embedData.R')
load(file = "./data/Fingers.rda")
```

```{r}
#| echo: false
#| results: 'asis'
cat(embedR("./assets/utils.R"))
cat(embedData(Fingers))
```


```{webr}
#| echo: false
#| output: true
#| warning: false
#| message: false

library(mosaic)
library(ggformula)
library(supernova)
options(digits = 3)

.webr_ready <- TRUE
set.seed(123)
```


:::{.nonincremental}
- **Modeling Data vs DGP**
- **Sampling Distribution**
- **Null Distribution of $b_1: (\mu_1-\mu_2)$**
- **Unlikely (unexpected) Events**
- **$\alpha$ vs. p-value**
- **Testing $b_1: \beta_1$**

:::

Perfect—this is where we can really shape the learning experience. I’ll give you **clean, ready-to-build slide structures** (works great for Quarto/RevealJS or PowerPoint), aligned with your quizzes and CourseKata flow.

I’ll keep:

* Minimal text per slide
* Strong narrative flow
* Built-in moments for questions/simulation

---

# 📘 **Session 1 Slides: Foundations of Inference**

## 🎯 Slide 1 — Title

**From Sampling to Inference**
*Subtitle: How do we decide if a result is “real”?*

---

## 🧠 Slide 2 — Big Question

> If we see a pattern in data…
> **Is it real or just random chance?**

---

## 🔁 Slide 3 — Recall: Sampling Variability

* Different samples → different statistics
* Statistics **vary naturally**

👉 Key idea:

> Variation is normal, not suspicious

---

## 📊 Slide 4 — Sampling Distribution

* Distribution of a statistic across many samples
* Built by:

  * Re-sampling OR
  * Mathematical theory

---

## 🔗 Slide 5 — Bridge to Inference

> What if there is **no real effect**?

We can ask:

* What would statistics look like **just by chance**?

👉 This leads to…

---

## ⚖️ Slide 6 — Null Hypothesis

**Null ($H_0$):**

* No effect
* No difference
* No relationship

Examples:

* Mean difference = 0
* Slope = 0

---

## 📉 Slide 7 — Null Distribution

* Distribution of a statistic **assuming $H_0$ is true**

👉 Built by:

* Simulation (shuffle)
* Mathematical models (normal, t, etc.)

---

## 🎯 Slide 8 — Key Idea

> We compare what we **observed**
> to what we would expect **by chance**

---

## 📏 Slide 9 — Test Statistic

* A number summarizing the effect

  * Mean difference
  * Slope
  * etc.

👉 Question:

> Is it unusually large (or small)?

---

## 📊 Slide 10 — Visual Idea

Sketch:

* Bell curve (null distribution)
* Mark observed statistic

---

## 📉 Slide 11 — P-value

**Definition:**

> Probability of getting a result this extreme (or more)
> **if the null is true**

---

## 🧠 Slide 12 — Interpretation

* Small p-value → unusual under null
* Large p-value → consistent with null

👉 Not:

* “Probability null is true”

---

## 🚨 Slide 13 — Decision Rule

* If p-value < α (e.g., 0.05)
  → Reject $H_0$

---

## ⚠️ Slide 14 — Type I Error

Rejecting a **true** null

👉 False positive

---

## ⚠️ Slide 15 — Type II Error

Failing to reject a **false** null

👉 Missed detection

---

## 🔁 Slide 16 — Full Workflow

1. State null
2. Compute statistic
3. Find null distribution
4. Compute p-value
5. Decide

---

## 🧪 Slide 17 — One Full Example (IMPORTANT)

Use:

* Mean test OR slope test

Walk through:

* Null: slope = 0
* Observed slope
* Null distribution
* p-value
* Decision

👉 This is the anchor of the session

---

## 🔍 Slide 18 — Assumptions (Light Touch)

* Independence
* Approximate normality (or large n)

👉 Why:

> Needed for mathematical null distributions

---

## 🔄 Slide 19 — Simulation vs Mathematical

* Simulation = flexible, intuitive
* Mathematical = faster, standard

👉 Same logic underneath

---

## 🎯 Slide 20 — Takeaway

> All inference = comparing data to a world where nothing is happening

---

---

# 📗 **Session 2 Slides: Extending the Framework**

## 🎯 Slide 1 — Title

**One Logic, Many Tests**

---

## 🔁 Slide 2 — Review

* Null hypothesis
* Null distribution
* P-value

👉 Same process every time

---

## 🧠 Slide 3 — Big Idea

> Different tests = different statistics
> Same logic = always compare to null

---

## 📊 Slide 4 — Multiple Groups (ANOVA Idea)

* Compare more than 2 means

Null:

> All group means are equal

---

## ⚖️ Slide 5 — ANOVA Statistic

* Measures:

  * Between-group variability
  * Relative to within-group variability

👉 Big value → evidence against null

---

## 📈 Slide 6 — Regression Test

Null:

> Slope ($b_1$) = 0

👉 No relationship

---

## 📊 Slide 7 — Regression Logic

* Compare observed slope
* To slopes expected by chance

👉 Same workflow as Session 1

---

## 🧮 Slide 8 — Chi-Square (New Context)

* Two categorical variables

Null:

> Variables are independent

---

## 📉 Slide 9 — Chi-Square Idea

* Compare:

  * Observed counts
  * Expected counts under independence

---

## 📊 Slide 10 — Shape of χ² Distribution

* Not symmetric
* Right-skewed

👉 Important conceptual shift

---

## 🎯 Slide 11 — Unifying Slide (VERY IMPORTANT)

| Test       | Statistic  | Null          |
| ---------- | ---------- | ------------- |
| Mean       | difference | no difference |
| Regression | slope      | slope = 0     |
| ANOVA      | F          | equal means   |
| Chi-square | χ²         | independence  |

👉 Same logic across all

---

## 🚨 Slide 12 — Type I Error Reminder

* False positive

---

## ⚠️ Slide 13 — Type I Error Inflation

* More tests → more false positives

---

## 📊 Slide 14 — Example

If α = 0.05:

* 1 test → 5% chance false positive
* 20 tests → much higher chance

---

## 🧠 Slide 15 — Why This Matters

* Real-world research
* Multiple comparisons

---

## 🛠️ Slide 16 — Practical Thinking

* Be cautious with many tests
* Look for patterns, not isolated results

---

## 🔁 Slide 17 — Full Circle

> Same 5-step process:

1. Null
2. Statistic
3. Null distribution
4. P-value
5. Decision

---

## 🎯 Slide 18 — Final Takeaway

> Inference is not about formulas
> It’s about comparing reality to randomness

---

---

# 💡 Optional Add-Ons (Highly Recommended)

### In-class activities

* **Session 1:** simulate null distribution (shuffle slope)
* **Session 2:** match test → null → statistic activity

