---
title: "CourseKata Chapter 10"
subtitle: "Logic of Inference"
author: "Mansour Abdoli, PhD"
format:
  live-revealjs:
    slide-number: true
    incremental: true
    chalkboard: true
    preview-links: true
    code-overflow: wrap
    footer: "CourseKata Ch. 10"
engine: knitr
execute:
  echo: true
  warning: false
  message: false
output-files: CK-ch10-complete.html
---

## Overview / Goals

{{< include ./_extensions/r-wasm/live/_knitr.qmd >}}

```{r}
#| echo: false
knitr::opts_chunk$set(echo=FALSE)
suppressPackageStartupMessages({
  library(mosaic)
  library(ggformula)
  library(supernova)
})
def.par <- par(no.readonly = TRUE)
on.exit(par(def.par))

source('./assets/utils.R')
source('./assets/embedR.R')
source('./data/embedData.R')
load(file = "./data/Fingers.rda")
```

```{r}
#| echo: false
#| results: 'asis'
cat(embedR("./assets/utils.R"))
cat(embedData(Fingers))
```


```{webr}
#| echo: false
#| output: true
#| warning: false
#| message: false

library(mosaic)
library(ggformula)
library(supernova)
options(digits = 3)

.webr_ready <- TRUE
set.seed(123)
```


:::{.nonincremental}
- **Modeling Data vs DGP**
- **Sampling Distribution**
- **Null Distribution of $b_1: (\mu_1-\mu_2)$**
- **Unlikely (unexpected) Events**
- **$\alpha$ vs. p-value**
- **Testing $b_1: \beta_1$**

:::

# Inference vs Fit
## Modeling Data vs DGP
::: {.columns}
::: {.column}
Modeling Data (Fit)

- Observed:
  - DATA = Model + Error
  - More complex, Better Fit
    - Fit: PRE
:::
::: {.column}
Modeling DGP (Inference)

- Unknown:
  - Hypothesis (Claim)
  - Null Hypothesis
    - Chance Model
:::
:::
<!-- end columns -->

- How good a model estimates DGP?
  - not sure, but it is not by chance => **Test of Hypothesis**
  - useful for prediction => **Cross-validation**

## Models
- Quantitative Response: $Y = Model + Error$
  - Categorical Explanatory $\Rightarrow b_0=\bar x_1$, $b_1=\bar x_2-\bar x_1$
  - Quantitative Explanatory $\Rightarrow b_0=\hat y_{x=0}$, $b_1=\text{slope}$
  
- Models have assumptions:
  - We focus on the parameter(s): $\beta_0$ or $\beta_1$

## Distributions
- Population Distribution (DGP)
  - Empty Model: $\beta_0 + \varepsilon$
  - Fixed shape, center and spread
  
- Sample Distribution
  - Data: $\bar x + e$
  - Everything may change per sample
  
- Distribution of a Sample Statistics
  - Sample Mean: $\beta_0 + \epsilon$
  - May change per sample size.

# Quantitative ~ Categorical
## Null Distribution of $b_1: (\mu_1-\mu_2)$
- Null: Empty Model
- $Y$ is independent of $X$
- Simulate Null Distribution of $\bar y$
  - Shuffle $Y$
  - Compute $b_1=\mu_2-\mu_1$
  - Repeat many times
- Visualize with a histogram
  
## Null Distribution Example
::: {.columns}
::: {.column}

```{webr}
breath<- read.csv('https://docs.google.com/spreadsheets/d/1y_imT-n_IyONZyvsFsGbUdRVYbJRxol-ODNPEGYe4_w/export?format=csv')
breath$swimmer <- factor(breath$swim_ability=="strong", 0:1, c('No', 'Yes'))
null_slopes <- do(500) * 
  b1(shuffle(breath_time) ~ swimmer, data = breath)
head(null_slopes)
```
:::
::: {.column}
```{webr}
#| fig-width: 5
#| fig-height: 4
#| warning: FALSE
gf_histogram(~swimmerYes, data = null_slopes) |>
  gf_point(0~b1(breath_time ~ swimmer, data = breath),
    color='red', size=2)

```
:::
:::

## Unlikely (unexpected) Events
- Observed Statistics far from Expected value
  - Distance in not universal
    - Use standardized value
    - Use Probability
- If **tail probability** is low,
  - observed statistic is unlikely

## Example
```{webr}
obs_b1<- b1(breath_time ~ swimmer, data = breath)
cat('Observed b1: ', obs_b1)
cat('Prop of b1>obs_b1: ', mean(null_slopes$swimmerYes>obs_b1))
gf_histogram(~swimmerYes, data = null_slopes, fill=~swimmerYes>obs_b1, bins=35) |>
  gf_point(0~obs_b1,
    color='red', size=2)

```

## $\alpha$ vs. $p$-value
How far is **far**?

- How low is an **unusual probability**?
  - Depends! But typically 5%.
  - We call this **Significance Level** ($\alpha$)

- Far: 
  - $p$-value$<  \alpha$
  - Observed statistic in $\alpha$ portion of tail(s) 

## Testing $b_1: (\mu_1-\mu_2)$
- Null Hypothesis: Empty Model DGP
  - Reject this if observed statistics in unusually far.

- In what direction?
  - The Alternative:
    - $\beta_1>0$: right tail
    - $\beta_1<0$: Left tail
    - $\beta_1\ne0$: both tails 
  
## Example: One-Tail Tests
:::{.columns}
:::{.column}
$H_a: \beta_1>0$
```{webr}
#| fig-width: 5
#| figfig-height: 3
gf_histogram(~swimmerYes, data = null_slopes, 
             fill=~swimmerYes>obs_b1, bins=35) |>
  gf_point(0~obs_b1, color='red', size=2) |>
  gf_labs(title=sprintf('p-value=%g', mean(null_slopes$swimmerYes>obs_b1)))

```
:::
:::{.column}
$H_a: \beta_1<0$
```{webr}
#| fig-width: 5
#| figfig-height: 3
gf_histogram(~swimmerYes, data = null_slopes, 
             fill=~swimmerYes<obs_b1, bins=35) |>
  gf_point(0~obs_b1, color='red', size=2) |>
  gf_labs(title=sprintf('p-value=%g', mean(null_slopes$swimmerYes<obs_b1)))

```
:::

## Example: Two-Tail Tests
:::{.columns}
:::{.column}
$H_a: \beta_1\ne 0$
```{webr}
#| fig-width: 5
#| figfig-height: 3
gf_histogram(~swimmerYes, data = null_slopes, 
             fill=~abs(swimmerYes)>abs(obs_b1), bins=35) |>
  gf_point(0~obs_b1, color='red', size=2) |>
  gf_labs(title=sprintf('p-value=%g',
                        mean(abs(null_slopes$swimmerYes)>abs(obs_b1))))

```
:::
:::{.column}
$F$ p-value
```{webr}
#| fig-width: 5
#| figfig-height: 3
supernova(lm(breath_time~swimmer, data = breath))

```
:::

## $F$ and $t$ Distributions
$F$ statistics in ANOVA:
  - Follows $F(df_{num}, df_{denum})$

:::{.columns}
:::{.column}
```{webr}
#| fig-width: 5
#| figfig-height: 3
supernova(lm(breath_time~swimmer, data = breath))

```
:::
:::{.column}
$F$ p-value
```{webr}
#| fig-width: 5
#| figfig-height: 3
df1<- 1; df2<- 41;
gf_dist("density", dist = "f", df1 = df1, df2 = df2, color = "blue") |>
  gf_ribbon(
    ymin = 0,
    ymax = ~df(x, df1, df2),
    data = data.frame(x = seq(3.651, 6, length.out = 200)),
    fill = "red", alpha = 0.4
  )
```
:::

# Quantitative ~ Quantitative
## Testing $b_1: \beta_1$
